---
title: "Chapter 1 — Did It Actually Work?"
subtitle: "Almaz, the Health Extension Workers, and the three traps"
---

## The mystery

Between 2000 and 2016, maternal mortality in Ethiopia fell from **673 to 412 deaths per 100,000 live births** *(DHS Ethiopia 2000; 2016)*.

Over the same period, Ethiopia trained about **38,000 Health Extension Workers** *(WHO 2013)*.

But facility delivery rates only rose from **5%** in 2005 to **26%** in 2016 *(DHS Ethiopia 2005; 2016)*. Most births were still happening at home during the years when mortality was falling fastest.

So what was actually working?

::: {.callout-note}
**The one thing to remember from this chapter:** Two things happening together is not the same as one causing the other — and figuring out which is which is the whole job of causal analysis.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch01-intro/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch01-intro/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — draw Almaz's picture

In the lecture, we drew Almaz's situation as a small diagram with four boxes — government funding at the top, the Health Extension Worker and the new road as siblings underneath, and the baby's survival at the bottom. In this workshop you'll draw that same diagram in R, then ask R one specific question: *if I add the road to my picture, does the apparent effect of the HEW change?*

You don't need to install anything yet. Read the code first. Run it later if you want.

### Setting up

We use a small package called `dagitty` for drawing causal diagrams, and `ggdag` for making them look nice on screen.

```{r}
#| label: ch01-setup
# Load the two packages we need.
# dagitty handles the math of causal diagrams.
# ggdag turns them into pictures.
library(dagitty)
library(ggdag)
```

### Almaz's picture, in code

This is the diagram from the lecture. Each line is one arrow.

```{r}
#| label: ch01-almaz-dag
# Build the diagram.
# Read each arrow as "this one influences that one".
almaz_dag <- dagitty('dag {
  Funding   -> HEW
  Funding   -> Road
  HEW       -> BabySurvives
  Road      -> BabySurvives
}')

# Tell ggdag where to put each box on the page.
coordinates(almaz_dag) <- list(
  x = c(Funding = 1, HEW = 0, Road = 2, BabySurvives = 1),
  y = c(Funding = 2, HEW = 1, Road = 1, BabySurvives = 0)
)

# Draw it.
ggdag(almaz_dag, text_size = 4) + theme_dag()
```

What you should see: a four-box picture. Funding sits at the top. Two arrows come down from it to the HEW box on the left and the Road box on the right. Then both of those have arrows pointing down to BabySurvives at the bottom.

This picture makes one thing visible that the words alone don't: **the HEW and the road share a parent.** They came in the same package. That's the whole reason it's hard to tell which one mattered.

### The one specific question

Now we ask `dagitty` the question from the lecture: *if we want to estimate the effect of the HEW on whether the baby survives, what else do we need to take into account?*

```{r}
#| label: ch01-adjustment-set
# Ask: to estimate HEW -> BabySurvives, what do we need to control for?
adjustmentSets(almaz_dag,
               exposure  = "HEW",
               outcome   = "BabySurvives")
```

R will tell you: **Road**.

That's the answer. If you want to know whether the HEW actually made a difference, you cannot just look at villages with HEWs versus villages without. You have to compare them at the same level of road access — because the road is doing some of the work, and the HEW is doing some of the work, and right now you can't tell how much of each.

This is the whole game we're going to play for the rest of the course: draw the picture, then let the picture tell you which questions are answerable and which ones aren't.

## Try it yourself

A neighbouring district to Almaz's launches a free SMS reminder service for pregnant women in 2018. By 2020, antenatal visit rates in that district have risen by 12 percentage points. The district health officer wants more funding to expand it.

Before you write your funding recommendation, draw the picture. What boxes would you put on it? What arrows? What other thing might have changed in that district between 2018 and 2020 — and where does *it* go on the picture?

We'll come back to this exact scenario in Chapter 2.
