---
title: "Chapter 1: Introduction to Causal Analysis"
subtitle: "Situational assessment, managerial intervention, and why correlation does not equal causation"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Distinguish between **situational assessment** ("What is happening?") and **managerial intervention** ("What should I do?")
- Explain why **correlation does not imply causation** using real maternal and newborn health examples from Ethiopia
- Identify the three classic causal thinking traps: **confounding**, **reverse causation**, and **selection bias**
- Describe the basic building blocks of a **causal model** -- nodes, arrows, and directed acyclic graphs (DAGs)
- Explain why causal reasoning matters for designing health programs that actually save lives
:::

## Key Concepts

::: {.key-concept}
#### Situational Assessment -- "What Is Happening?"

Every decision starts with understanding the current state of the world. A doctor examines a patient before prescribing medicine. A mechanic listens to the engine before opening the hood. A health program manager in Ethiopia asks: "Why is neonatal mortality not falling in Somali region -- is it a workforce problem, a supply chain problem, or something else entirely?" Situational assessment means using evidence to figure out which variables are in which states, and *why*. You cannot fix a problem you do not understand.
:::

::: {.key-concept .green}
#### Managerial Intervention -- "What Should I Do?"

Once you understand the situation, you face a choice: which levers should you pull? Resources are always limited -- money, time, people, equipment. Spending on training means *not* spending on equipment. Building a data system means *not* hiring more midwives. Every intervention involves trade-offs, and choosing wisely requires understanding what causes what. Ethiopia's Health Extension Program invested across three pillars -- People (40,000+ community health workers), Products (essential medicines and equipment), and Systems (DHIS2 and referral networks). The causal question is: how do these three pillars combine to produce better outcomes? Do they work independently, or do they need each other?
:::

::: {.key-concept .red}
#### The Three Causal Thinking Traps

| Trap | What Goes Wrong | Ethiopia Example |
|---|---|---|
| **Confounding** | A hidden third variable drives both X and Y | Disease burden drives both health worker deployment and mortality -- regions with the most workers have the highest deaths |
| **Reverse Causation** | Y causes X, not X causes Y | High mortality attracts international funding, not the other way around |
| **Selection Bias** | Your data sample is not representative | Only well-resourced hospitals report to DHIS2, so equipment appears unrelated to survival |

All three traps create patterns in data that look like causal effects but are not. The antidote is explicit causal modeling -- drawing a picture of what causes what *before* you analyze the data.
:::

::: {.key-concept .teal}
#### Directed Acyclic Graphs (DAGs)

A causal model is just a diagram that shows what causes what. It has three simple ingredients: **nodes** (the things you care about, drawn as circles or shapes), **arrows** (causal links -- "A causes B" is drawn as A to B), and **signs** ([+] or [-] telling you the direction of effect). A DAG is a causal diagram where all arrows point in one direction and there are no loops -- you can never follow the arrows from any node back to itself. "Directed" means arrows have a direction. "Acyclic" means no circles. "Graph" is just the math word for a diagram of connected nodes. The no-loops rule matters because loops make it impossible to figure out what causes what.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch01-intro/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch01-intro/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This chapter introduces the core R packages for causal analysis: `dagitty` for defining and analyzing DAGs, `ggdag` for drawing them, and `ggplot2` for general plotting. By the end of this section you will have built three different causal diagrams in code.

### Setting Up the Environment

Before we draw any diagrams, we need to load our tools. Think of these packages as specialized toolboxes: `dagitty` lets you write down a causal diagram and ask questions about it, `ggdag` turns the code into a visual picture, and `ggplot2` powers the drawing engine behind the scenes.

```{r}
#| label: setup
#| echo: true

# Install packages the first time you use them (remove the # to run):
# install.packages(c("dagitty", "ggdag", "ggplot2"))

# Load the packages into our R session
library(dagitty)    # Define and analyze causal diagrams (DAGs)
library(ggdag)      # Draw pretty pictures of those DAGs
library(ggplot2)    # General-purpose plotting library
```

### A Simple 3-Node Chain

Let's start with the simplest useful causal model for a health program. We define three nodes -- Investment, Coverage, and Mortality -- connected by two arrows. Investment increases coverage, and higher coverage reduces mortality. Notice that investment does not directly reduce mortality; it works *through* coverage.

```{r}
#| label: simple-dag
#| echo: true
#| fig-width: 8
#| fig-height: 4

# Define a 3-node DAG
# pos="x,y" sets the position on the diagram
mnh_simple <- dagitty('dag {
  Investment [pos="0,0"]
  Coverage   [pos="1,0"]
  NMR        [pos="2,0"]

  Investment -> Coverage
  Coverage -> NMR
}')

# Draw the diagram
# theme_dag() removes gridlines and axes for a clean look
ggdag(mnh_simple) +
  theme_dag() +
  labs(title = "Simple MNH Causal Model",
       subtitle = "Investment -> Coverage -> Neonatal Mortality Rate")
```

This tiny model already tells us something useful: if investment goes up but coverage does *not* go up, the money is leaking somewhere in between. The chain helps you find where things break down.

### A Confounded Model

Now let's add Complication Severity as a confounder -- the hidden variable that creates a misleading correlation between hospital delivery and mortality. This is the classic Ethiopia puzzle: regions where the most health workers are deployed tend to have the highest mortality rates.

```{r}
#| label: confounded-dag
#| echo: true
#| fig-width: 9
#| fig-height: 5

# Hospital delivery and mortality: confounded by complication severity
confounded_dag <- dagitty('dag {
  Complication_Severity [pos="1,0"]
  Hospital_Delivery     [pos="0,1"]
  Neonatal_Mortality    [pos="2,1"]

  Complication_Severity -> Hospital_Delivery
  Complication_Severity -> Neonatal_Mortality
  Hospital_Delivery -> Neonatal_Mortality
}')

# Color-code by role: green = exposure, red = outcome, amber = confounder
ggdag_status(confounded_dag,
             exposure = "Hospital_Delivery",
             outcome = "Neonatal_Mortality") +
  theme_dag() +
  scale_color_manual(values = c("exposure" = "#83BD00",
                                "outcome" = "#E24A3F",
                                "other" = "#F8A623")) +
  labs(title = "Confounding in Hospital Delivery Data",
       subtitle = "Complication severity drives both delivery location and mortality")
```

Read the picture: Complication Severity sits at the top, sending arrows down to *both* Hospital Delivery and Neonatal Mortality. That is the fork structure -- the classic confounding pattern. Without accounting for severity, you would wrongly conclude that hospital delivery increases mortality.

### Finding Adjustment Sets

Here is the powerful part. Once we have drawn the diagram, R can tell us *exactly* which variables we need to control for to get the true causal effect. The `adjustmentSets()` function does this automatically.

```{r}
#| label: adjustment-sets
#| echo: true

# Ask dagitty: "What do I need to control for to estimate the
# causal effect of Hospital_Delivery on Neonatal_Mortality?"
adjustmentSets(confounded_dag,
               exposure = "Hospital_Delivery",
               outcome = "Neonatal_Mortality")
# Returns: { Complication_Severity }
# Translation: control for severity, and the confounding goes away
```

In plain English: if you compare patients who have the *same* complication severity, then hospital delivery is associated with *lower* mortality. The confounding disappears once you account for why patients ended up at the hospital in the first place.

### Ethiopia's Three-Pillar Model

Let's build a richer model that captures Ethiopia's three-pillar approach -- People, Products, and Systems -- all flowing through Quality of Care, with Disease Burden as a confounder affecting both mortality outcomes.

```{r}
#| label: rich-dag
#| echo: true
#| fig-width: 10
#| fig-height: 6

# Ethiopia's three-pillar theory of change as a DAG
ethiopia_toc <- dagitty('dag {
  People          [pos="1,0"]
  Products        [pos="1,1"]
  Systems         [pos="1,2"]
  Quality         [pos="2,1"]
  NMR             [pos="3,0.5"]
  MMR             [pos="3,1.5"]
  Disease_Burden  [pos="2,2.5"]

  People   -> Quality
  Products -> Quality
  Systems  -> Quality
  Quality  -> NMR
  Quality  -> MMR
  Disease_Burden -> NMR
  Disease_Burden -> MMR
}')

# Find the adjustment set for estimating People -> NMR
adjustmentSets(ethiopia_toc,
               exposure = "People",
               outcome = "NMR")

# Draw the full model
ggdag(ethiopia_toc) +
  theme_dag() +
  labs(title = "Ethiopia MNH: Three-Pillar Theory of Change",
       subtitle = "People, Products, and Systems -> Quality -> Mortality")
```

This model says all three pillars flow into Quality of Care. Quality then reduces both NMR and MMR. Disease Burden affects the outcomes directly -- a region with severe malaria burden will have higher mortality even with good care. A trained midwife without a PPH kit cannot stop a hemorrhage. A PPH kit without a trained midwife is just a box on a shelf. People, Products, and Systems are *complements* -- they multiply each other's effect.

## Key Takeaways

::: {.key-takeaway}
**Causal thinking starts with two questions:** "What is happening?" (situational assessment) and "What should I do?" (managerial intervention). You cannot answer the second well without answering the first.

**Correlation does not equal causation.** Three traps -- confounding, reverse causation, and selection bias -- all create misleading patterns in data. Ethiopia's 40,000 Health Extension Workers are not causing deaths. Disease burden is confounding the relationship.

**A causal model (DAG)** makes your assumptions visible, testable, and actionable. Draw the diagram. Ask R what to control for. Analyze the data correctly. Make better decisions.
:::

## Looking Ahead

Next in **Chapter 2 -- Qualitative Causal Models**, we learn the formal language of causal diagrams:

- Three types of nodes: probabilistic (ovals), decision (rectangles), and objective (hexagons)
- Signed directed links with [+] and [-] telling you which direction things move
- The three fundamental triplet structures: **chains**, **forks**, and **colliders** -- and why colliders behave opposite to everything else
- **d-Separation** -- the rule that tells you when information flows between variables and when it stops
