---
title: "Chapter 2 — Drawing the Picture"
subtitle: "Beatrice, the Animatrices, and the grammar of causal diagrams"
---

## The mystery

Rwanda's neonatal mortality rate fell from **41 to 19 per 1,000 live births** between 2000 and 2019 *(DHS Rwanda 2000; 2019/20)*. The front line was about **45,000 community health workers** — almost all women, almost all volunteers, with six days of training each *(Rwanda Biomedical Centre, 2020)*.

How can volunteers with six days of training help drive one of the fastest declines on record? The answer is in the *picture*, not the training.

::: {.callout-note}
**The one thing to remember from this chapter:** A causal diagram is just three node types, signed arrows, and three shapes — and once you know that grammar, you can draw a picture for any programme you encounter.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch02-qualitative-models/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch02-qualitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — build Beatrice's diagram, then check the shapes

In the lecture, we built Beatrice's full picture: a fork at the top (the government CHW programme spawning both her training and the new ambulance route), a chain running through her visits and the mother's recognition of danger signs, and a collider at the survival outcome. In this workshop you'll build that diagram in R and ask `dagitty` to *find* the chains, forks, and colliders for you — so you don't have to spot them by eye.

### Setting up

```{r}
#| label: ch02-setup
# Same two packages as Chapter 1.
library(dagitty)
library(ggdag)
```

### Beatrice's full picture, in code

```{r}
#| label: ch02-beatrice-dag
# Build the diagram exactly as it appeared in the lecture.
# Read each line as "this one influences that one".
beatrice_dag <- dagitty('dag {
  CHW_Programme   -> Training
  CHW_Programme   -> Ambulance
  Training        -> Visits
  Visits          -> DangerSigns
  DangerSigns     -> NewbornSurvives
  Ambulance       -> NewbornSurvives
  Distance        -> NewbornSurvives
  Distance        -> Visits
  Poverty         -> Visits
  Poverty         -> NewbornSurvives
}')

coordinates(beatrice_dag) <- list(
  x = c(CHW_Programme = 1, Training = 0, Ambulance = 2,
        Visits = 0, DangerSigns = 1, NewbornSurvives = 2,
        Distance = 2, Poverty = -1),
  y = c(CHW_Programme = 3, Training = 2, Ambulance = 2,
        Visits = 1, DangerSigns = 1, NewbornSurvives = 0,
        Distance = 2, Poverty = 1)
)

ggdag(beatrice_dag, text_size = 3) + theme_dag()
```

### Let R find the shapes for you

`dagitty` knows the grammar from Chapter 2. You can ask it to list the chains, forks, and colliders without spotting them by eye.

```{r}
#| label: ch02-find-shapes
# All the paths between the action (Visits) and the outcome.
# dagitty marks each path as 'open' (the influence flows) or 'blocked'.
paths(beatrice_dag, from = "Visits", to = "NewbornSurvives")
```

What you should see: at least one open path going `Visits -> DangerSigns -> NewbornSurvives` (that's the chain), and at least one path involving `Poverty` (which is a fork — Poverty is the shared parent of Visits and NewbornSurvives).

Now let's ask the same `adjustmentSets` question we asked in Chapter 1 — but with a richer picture.

```{r}
#| label: ch02-adjustment
# What do we need to adjust for to estimate Visits -> NewbornSurvives?
adjustmentSets(beatrice_dag,
               exposure = "Visits",
               outcome  = "NewbornSurvives")
```

R will tell you which variables you'd need to control for. Notice: it does **not** include `DangerSigns`, because `DangerSigns` is on the chain *between* Visits and the outcome — adjusting for it would block the very path you're trying to measure. The grammar from the lecture is doing real work here.

## Try it yourself

A new programme will train CHWs in a neighbouring district to use a smartphone app for danger-sign screening. Before you fund it, draw the picture. Where does *the smartphone* go? Where does *phone signal* go? Where does *literacy* go? Where does *the mother's trust in technology* go?

In Chapter 3, we'll turn this exact exercise into a method — building a causal picture by interviewing the people who run the programme.
