---
title: "Chapter 2: Qualitative Causal Models"
subtitle: "Nodes, directed links, triplet structures, and d-separation"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Identify the **three main node types** in a causal diagram -- probabilistic, decision, and objective -- and know when to use each
- Assign **signed directed links** [+]/[-] to represent how one variable pushes another up or down
- Recognize the three fundamental **triplet structures**: chain, fork, and collider
- Apply **d-separation** rules to determine when information flows between variables and when it stops
- Build a complete **qualitative DAG** for Rwanda's maternal and newborn health system using `dagitty` in R
:::

## Key Concepts

::: {.key-concept}
#### Three Node Types

A qualitative causal model uses three main types of nodes. **Probabilistic nodes** (drawn as ovals) represent variables whose value is uncertain -- things the world throws at you, like quality of care, care-seeking behavior, or workforce competency. **Decision nodes** (drawn as rectangles) represent choices you make -- the levers you can pull, like how much to invest in training or whether to expand Rwanda's Mutuelle de Sante insurance. **Objective nodes** (drawn as hexagons) represent the goals you are trying to achieve -- like reducing neonatal mortality. Arrows flow *into* objective nodes, never *out*. They are the scoreboard. There is also a fourth type -- function nodes -- for variables calculated exactly from their parents (like Total Spending = Training + Equipment + Systems), but we will not focus on those here.
:::

::: {.key-concept .green}
#### Signed Directed Links

An arrow from node A to node B means that changing A can change B. The sign tells you the direction. **[+]** means "more A leads to more B" -- A pushes B up. **[-]** means "more A leads to less B" -- A pushes B down. For example: CHW Training Investment --[+]--> Workforce Competency --[-]--> Neonatal Mortality. Training *increases* competency, and higher competency *decreases* mortality. When you follow a path through multiple arrows, you multiply the signs: [+] times [+] = [+], [-] times [-] = [+], and [+] times [-] = [-]. Same rule as multiplying positive and negative numbers. So the overall sign of Training to Mortality is [+] times [-] = [-]. Training reduces mortality.
:::

::: {.key-concept .orange}
#### The Three Triplet Structures

Every DAG -- no matter how large -- is built from combinations of just three patterns. In a **chain** (A --> B --> C), B is a mediator that passes information from A to C. Think: Rain --> Wet Roads --> Car Accidents. If you already know the roads are wet, learning it rained tells you nothing new about accident risk. Conditioning on B *blocks* the flow. In a **fork** (A <-- B --> C), B is a common cause of both A and C. Think: Ice Cream Sales <-- Summer Heat --> Drowning Deaths. Ice cream does not cause drowning; hot weather drives both. Conditioning on B *removes* the spurious correlation. In a **collider** (A --> B <-- C), two arrows crash into B. Think: High Grades --> University Admission <-- Athletic Talent. Among admitted students, non-athletes must be academic stars. This path is *blocked* by default, but conditioning on B *opens* it. Colliders are the rebel -- they behave opposite to chains and forks.
:::

::: {.key-concept .red}
#### Information Flow Rules (d-Separation)

Two nodes X and Y are **d-separated** given a conditioning set Z if every path between them is blocked. A path is blocked if it contains: (1) a chain or fork whose middle node is conditioned on, OR (2) a collider whose middle node (and all its descendants) is *not* conditioned on. d-Separation means conditional independence -- knowing X tells you nothing about Y once you account for Z. These are testable predictions of the model. If your data violates an implied independence, the model must be revised. The practical shortcut: walk along each path, check each node. Chain/fork middle conditioned on? Blocked. Collider *not* conditioned on? Blocked. One blocked node is enough to block the whole path.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch02-qualitative-models/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch02-qualitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This chapter builds Rwanda's MNH qualitative DAG step by step in R. We use `dagitty` for DAG specification and d-separation testing, and `ggdag` for visualization. Rwanda provides a compelling case: its Mutuelle de Sante universal insurance program covers approximately 85% of the population, and its community health worker network spans all 30 districts -- yet neonatal mortality (~16 per 1,000 live births, WHO) has been harder to reduce than under-5 mortality.

### Setting Up

```{r}
#| label: setup
#| echo: true

# Install packages (run once -- uncomment the line below)
# install.packages(c("dagitty", "ggdag", "ggplot2"))

library(dagitty)    # Define and analyze DAGs
library(ggdag)      # Visualize DAGs using ggplot2
library(ggplot2)    # General plotting
```

### Step 1: Define Rwanda's DAG

We specify a 7-node qualitative DAG with three decision nodes (the levers Rwanda's health leadership can pull), three probabilistic nodes (mediators whose values are uncertain), and one objective node (neonatal mortality -- the outcome we want to reduce). The `pos` coordinates control the layout on the diagram.

```{r}
#| label: define-dag
#| echo: true

# Rwanda MNH Qualitative DAG
# 3 decision nodes, 3 probabilistic mediators, 1 objective
rwanda_dag <- dagitty('dag {
  # Decision nodes -- things Rwanda's health system controls
  CHW_Training       [pos="0,0"]
  Equipment_Procure  [pos="0,1"]
  Insurance_Policy   [pos="0,2"]

  # Probabilistic nodes -- uncertain, shaped by upstream causes
  Competency         [pos="1,0"]
  Equip_Avail        [pos="1,1"]
  Care_Seeking       [pos="1,2"]
  Quality            [pos="2,1"]

  # Objective node -- the outcome we want to reduce
  Neonatal_Mortality [pos="3,1"]

  # Causal arrows (each one means "changing the parent can change the child")
  CHW_Training      -> Competency
  Equipment_Procure -> Equip_Avail
  Insurance_Policy  -> Care_Seeking
  Competency        -> Quality
  Equip_Avail       -> Quality
  Care_Seeking      -> Quality
  Quality           -> Neonatal_Mortality
}')

# Verify it has no cycles -- a valid DAG
isAcyclic(rwanda_dag)   # Should return TRUE
```

### Step 2: Visualize the DAG

The `ggdag()` function turns our code into a picture. The `ggdag_status()` variant color-codes nodes by their role -- exposure (the cause we are studying) and outcome (what we are measuring). This makes it easy to see the causal pathway at a glance.

```{r}
#| label: visualize-dag
#| echo: true
#| fig-width: 10
#| fig-height: 5

# Basic visualization
ggdag(rwanda_dag) +
  theme_dag() +
  labs(title    = "Rwanda MNH: Qualitative Causal Model",
       subtitle = "7-node DAG for neonatal health investment")

# Color-coded by exposure and outcome
ggdag_status(rwanda_dag,
             exposure = "CHW_Training",
             outcome  = "Neonatal_Mortality") +
  theme_dag() +
  labs(title = "Exposure: CHW Training | Outcome: Neonatal Mortality")
```

Read the diagram left to right: decisions (what Rwanda controls) flow through mediators (how the decisions work) into Quality of Care (the bottleneck where everything converges), and then into Neonatal Mortality (the outcome). Notice that Quality is a *collider* -- Competency, Equipment Availability, and Care-Seeking all send arrows into it.

### Step 3: Test d-Separation

The `dseparated()` function lets us check whether two variables are conditionally independent given a set of other variables. Each test verifies a prediction the DAG makes about the data.

```{r}
#| label: dsep-tests
#| echo: true

# Test 1: Is CHW_Training independent of Mortality given Quality?
# This is a chain: Training -> Competency -> Quality -> Mortality
# Conditioning on Quality should block the path
dseparated(rwanda_dag,
           "CHW_Training", "Neonatal_Mortality", "Quality")
# Expected: TRUE -- chain blocked at Quality

# Test 2: Is Competency independent of Equip_Avail? (no conditioning)
# These are separate branches with no connecting path
dseparated(rwanda_dag,
           "Competency", "Equip_Avail", c())
# Expected: TRUE -- no connecting path

# Test 3: Is Competency independent of Equip_Avail given Quality?
# Quality is a collider! Conditioning on it opens the path
dseparated(rwanda_dag,
           "Competency", "Equip_Avail", "Quality")
# Expected: FALSE -- collider bias opens the path!

# Test 4: Is Insurance_Policy independent of CHW_Training?
# Both are root nodes with no common parent
dseparated(rwanda_dag,
           "Insurance_Policy", "CHW_Training", c())
# Expected: TRUE -- separate decisions, no shared cause
```

Test 3 is the most important result. It shows the collider effect in action: Competency and Equipment Availability are completely independent by default (Test 2), but the moment you condition on Quality -- for example, by only studying high-quality facilities -- a spurious link appears. Among good-quality facilities, if staffing is low, equipment *must* be excellent. Each cause "explains away" the other.

### Step 4: Paths, Adjustment Sets, and Testable Predictions

The `paths()` function lists every route from one node to another. The `adjustmentSets()` function tells you what you must control for in your data analysis. And `impliedConditionalIndependencies()` generates every testable prediction the model makes.

```{r}
#| label: paths-and-adjustments
#| echo: true

# All paths from CHW_Training to Neonatal_Mortality
paths(rwanda_dag, from = "CHW_Training", to = "Neonatal_Mortality")

# What must we control for to estimate the causal effect of
# CHW_Training on Neonatal_Mortality?
adjustmentSets(rwanda_dag,
               exposure = "CHW_Training",
               outcome  = "Neonatal_Mortality")
# Returns {} -- no confounders! CHW_Training is a root node
# with no backdoor paths to close

# All testable predictions of the model
# If Rwanda's DHS or DHIS2 data violates any of these,
# the model needs revision
impliedConditionalIndependencies(rwanda_dag)
```

The empty adjustment set is significant. Because CHW_Training has no parents in this DAG, there are no backdoor paths to close. The causal effect can be estimated without adjustment -- *if* the DAG is correct. The implied conditional independencies give us a way to check: if the data shows that, say, Equipment Availability and Care-Seeking are correlated even without conditioning on Quality, then our model is missing a link (perhaps a shared "Total Health Budget" parent).

## Key Takeaways

::: {.key-takeaway}
**A qualitative causal model makes your assumptions visible, shareable, and testable.** Instead of vague claims like "training improves outcomes," you draw a specific diagram that everyone can challenge.

Every DAG is built from three triplets -- **chains**, **forks**, and **colliders** -- each with its own information flow rule. Chains and forks: conditioning on the middle *blocks* the flow. Colliders: conditioning on the middle *opens* the flow. Get this backwards and you introduce bias instead of removing it.

For Rwanda: three investment decisions flow through three mediators into Quality of Care, then Neonatal Mortality. The model tells us *where* to look -- the numbers (Chapter 4) will tell us *how much*.
:::

## Looking Ahead

Next in **Chapter 3 -- The MNH Diagnostic Case Study**, we put qualitative models to work on a real consulting engagement in Kenya:

- Building a DAG from stakeholder interviews through an iterative process -- starting with just 4 nodes and expanding round by round
- Interviewing a county health director, a senior midwife, and a community health worker -- each reveals variables the others cannot see
- Discovering the "transmission problem" -- why inputs (training, equipment) are not becoming outputs (quality, survival)
- Using root nodes and common causes to prioritize recommendations
