---
title: "Chapter 3 — Building the Picture from Scratch"
subtitle: "Faith, three interviews, and one shared picture"
---

## The mystery

Kenya devolved health to **47 counties** in 2013 *(Constitution of Kenya, Schedule 4)*. Some counties have cut maternal mortality dramatically. Others have barely moved *(Kenya Health Information System, county dashboards)*. The county dashboards can tell you *which* counties are improving — but not *why*.

To find out why, you have to go talk to people. The question is who, in what order, and what to do with what they tell you.

::: {.callout-note}
**The one thing to remember from this chapter:** The fastest way to build a useful causal picture is to interview the people who actually do the work — in three different groups, in a particular order — and draw their answers.
:::

## Main lecture

::: {.slide-embed}
<iframe src="../ch03-interview-case/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch03-interview-case/index.html" target="_blank">Open lecture slides full-screen &rarr;</a>
</div>
:::

## Tutorial / breakout session

The tutorial deck takes the same ideas and retells them as a story — built for a live breakout session, not a lecture.

::: {.slide-embed}
<iframe src="../ch03-interview-case-tutorial/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch03-interview-case-tutorial/index.html" target="_blank">Open tutorial slides full-screen &rarr;</a>
</div>
:::

## R workshop — assemble Faith's three interviews

In the lecture, Faith built her picture from three interviews — Mary the midwife, Lokol the husband, Esther the finance officer. Each one added boxes the others couldn't see. In this workshop you'll do the same thing in R: start with Mary's chain, add Lokol's permission box, add Esther's stockout box, and watch how the picture grows.

### Setting up

```{r}
#| label: ch03-setup
library(dagitty)
library(ggdag)
```

### Step 1 — Mary's picture (the bedside)

```{r}
#| label: ch03-mary
mary_dag <- dagitty('dag {
  Bleeding -> FamilyWaits
  FamilyWaits -> Motorbike
  Motorbike -> ArrivesAtFacility
  ArrivesAtFacility -> MotherDies
}')
ggdag(mary_dag, text_size = 3) + theme_dag()
```

A clean chain. Five boxes, four arrows. This is what Mary saw — and only what Mary saw.

### Step 2 — Lokol's correction (the household)

Lokol's interview reveals that "FamilyWaits" was actually two boxes — the deciding, and the permission from the head of family. Let's split it.

```{r}
#| label: ch03-lokol
mary_lokol_dag <- dagitty('dag {
  Bleeding -> FamilyDecides
  FamilyDecides -> Permission
  Permission -> Motorbike
  Motorbike -> ArrivesAtFacility
  ArrivesAtFacility -> MotherDies
}')
ggdag(mary_lokol_dag, text_size = 3) + theme_dag()
```

The chain grew by one box. The new box (`Permission`) sits where Mary's "waits" used to be — and it's the box Mary couldn't see, because she wasn't in the household.

### Step 3 — Esther's layer (the supply chain)

Esther's stockout has nothing to do with the bedside chain — it sits in a different part of the picture entirely, with its own arrow into the death-outcome box.

```{r}
#| label: ch03-full
faith_dag <- dagitty('dag {
  NationalSupplier -> CountyStockout
  CountyStockout -> DrugAtFacility
  DrugAtFacility -> MotherDies

  Bleeding -> FamilyDecides
  FamilyDecides -> Permission
  Permission -> Motorbike
  Motorbike -> ArrivesAtFacility
  ArrivesAtFacility -> MotherDies
}')

coordinates(faith_dag) <- list(
  x = c(NationalSupplier = 0, CountyStockout = 1, DrugAtFacility = 2,
        Bleeding = 0, FamilyDecides = 1, Permission = 2,
        Motorbike = 3, ArrivesAtFacility = 4, MotherDies = 5),
  y = c(NationalSupplier = 2, CountyStockout = 2, DrugAtFacility = 2,
        Bleeding = 0, FamilyDecides = 0, Permission = 0,
        Motorbike = 0, ArrivesAtFacility = 0, MotherDies = 1)
)

ggdag(faith_dag, text_size = 2.5) + theme_dag()
```

What you should see: two layers feeding into a single death-outcome. The bedside layer at the bottom. The supply-chain layer up top. Joined only at the very end. *(We use the death framing because Mary's interview was specifically about what causes deaths — and the picture should reflect what the source actually said.)*

### Asking the picture which arrows you'd need to test

```{r}
#| label: ch03-paths
# How many distinct paths run from the start of the bleeding chain to the
# death-outcome? Each one is a place a programme could intervene.
paths(faith_dag, from = "Bleeding", to = "MotherDies")
```

R will show you that there's only one path — but it has six links in it. Every one of those links is a place a programme could intervene to *break the chain* and save the life. And the picture tells you that *DrugAtFacility* is on a completely separate path, controlled by an entirely different set of people.

This is the practical payoff of the interview method: the finished picture *makes the intervention points visible*.

## Try it yourself

You have one week and a budget for petrol. You need to build a causal picture explaining why one ward in a Kenyan county has much higher under-5 mortality than its three neighbours. Who do you interview, in what order, and what is the opening question for each?

In Chapter 4, we'll take a picture like Faith's and start putting actual numbers on the arrows.
