---
title: "Chapter 4: Quantitative Causal Models"
subtitle: "Adding numbers to your causal diagrams with probability tables, Bayes' rule, and Bayesian networks"
---

## Learning Objectives

::: {.learning-objectives}
- Translate everyday observations into **natural frequencies** ("55 out of 100 facilities") and then into probabilities
- Build a **Conditional Probability Table (CPT)** -- a lookup table that answers "given the parents, how likely is each outcome?"
- Apply **Bayes' rule** using a tree diagram to flip from "what would I see if this were true?" to "given what I see, what is probably true?"
- Combine a DAG with CPTs to create a **Bayesian network** and query it in R using `bnlearn`
- Explain why the **Causal Markov Condition** keeps the math manageable even in large models
:::

## Key Concepts

::: {.key-concept}
#### Natural Frequencies and Conditional Probability

Probabilities are easier to grasp when you start with counts. Imagine 100 health facilities in Tanzania. About 40 have adequate staffing; 60 do not. Among the 40 with adequate staffing, 34 deliver good-quality care. Among the 60 without, only 12 do. Those raw counts -- 34 out of 40, 12 out of 60 -- are natural frequencies. Divide them to get conditional probabilities: P(Good Quality | Adequate Staffing) = 34/40 = 0.85. Natural frequencies make the numbers feel real before we ever write a formula.
:::

::: {.key-concept .green}
#### Lookup Tables (Conditional Probability Tables)

A CPT is just a lookup table. You find the row that matches the parent states, and you read off the probability of each child state. Every row must add up to 1 (something has to happen). For a node with two parents each having two states, the table has four rows -- one for every combination. CPTs are the building blocks that turn a qualitative DAG into a number-crunching machine. In Tanzania's context, a CPT might encode: "If staffing is adequate AND equipment is available, there is an 85% chance of good-quality care."
:::

::: {.key-concept .orange}
#### Bayes' Rule (Tree Diagram Version)

Bayes' rule lets you reason backward. You observe an outcome -- say, high neonatal mortality in a Tanzanian district -- and you want to know the most likely cause. The tree diagram makes this visual: draw branches for each possible cause (staffing adequate vs. low), then sub-branches for the outcome (high vs. low mortality). Multiply along each branch to get the joint probability, then normalize. The result? Your updated belief about staffing, given what you observed. Prior belief in, evidence applied, posterior belief out.
:::

::: {.key-concept .red}
#### Bayesian Networks

A Bayesian network = DAG + one CPT per node. That is it. The DAG tells you what causes what. The CPTs tell you how strong each link is. Together, they let you compute any probability you want, update beliefs when new evidence arrives, and compare interventions side by side. For Tanzania, a four-node network (Staffing, Equipment, Quality, Neonatal Mortality) can answer questions like: "If we fix staffing but not equipment, how much does mortality drop?"
:::

::: {.key-concept .teal}
#### The Causal Markov Condition

Here is the trick that makes Bayesian networks practical. Each variable is independent of its non-descendants once you know its parents. This means the full joint distribution -- which could have millions of entries -- factors into a small product of CPTs. Instead of specifying every possible combination of every variable, you only specify each node's relationship with its direct parents. The DAG structure does the heavy lifting.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch04-quantitative-models/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch04-quantitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

Time to build a Bayesian network from scratch in R. We will model a simplified version of Tanzania's neonatal health system: four nodes, three arrows, and one big question -- which investment saves the most lives?

### Define the DAG

We use `bnlearn` to specify the structure. The notation `[Node|Parent]` means "this node depends on that parent."

```{r}
#| label: dag-structure
#| echo: true

library(bnlearn)

# Four-node DAG for Tanzania neonatal health
# Staffing and Equipment both cause Quality; Quality causes NMR
dag <- model2network(
  "[Staffing][Equipment][Quality|Staffing:Equipment][NMR|Quality]"
)

# Check what we built
dag
nodes(dag)   # "Staffing" "Equipment" "Quality" "NMR"
arcs(dag)    # Three directed edges
```

### Fill in the CPTs

Each CPT is an array whose dimensions match the node's states and its parents' states. Root nodes (no parents) get simple marginal distributions.

```{r}
#| label: cpt-specification
#| echo: true

# Root nodes: simple probabilities based on Tanzania DHS data
cpt_staffing <- array(c(0.40, 0.60), dim = 2,
  dimnames = list(Staffing = c("Adequate", "Low")))

cpt_equipment <- array(c(0.35, 0.65), dim = 2,
  dimnames = list(Equipment = c("Available", "Unavailable")))

# Quality depends on both Staffing and Equipment (4 rows, each sums to 1)
cpt_quality <- array(
  c(0.85, 0.15,   # Adequate staffing + Available equipment
    0.50, 0.50,   # Low staffing + Available equipment
    0.55, 0.45,   # Adequate staffing + Unavailable equipment
    0.20, 0.80),  # Low staffing + Unavailable equipment
  dim = c(2, 2, 2),
  dimnames = list(
    Quality   = c("Good", "Poor"),
    Staffing  = c("Adequate", "Low"),
    Equipment = c("Available", "Unavailable")))

# NMR depends on Quality
cpt_nmr <- array(
  c(0.85, 0.15,   # Good quality -> 85% chance of low NMR
    0.40, 0.60),  # Poor quality -> 60% chance of high NMR
  dim = c(2, 2),
  dimnames = list(NMR = c("Low", "High"), Quality = c("Good", "Poor")))

# Assemble into a fitted Bayesian network
bn_tz <- custom.fit(dag, list(
  Staffing = cpt_staffing, Equipment = cpt_equipment,
  Quality = cpt_quality, NMR = cpt_nmr))
```

### Query the Network

Now the fun part. We can ask the network any conditional probability question using `cpquery()`.

```{r}
#| label: bn-queries
#| echo: true

# Question 1: What is P(NMR = High) overall in Tanzania?
cpquery(bn_tz, event = (NMR == "High"), evidence = TRUE, n = 1e6)

# Question 2: If staffing is low, how likely is high NMR?
cpquery(bn_tz,
  event    = (NMR == "High"),
  evidence = (Staffing == "Low"), n = 1e6)

# Question 3: Best case -- adequate staffing AND available equipment
cpquery(bn_tz,
  event    = (NMR == "High"),
  evidence = (Staffing == "Adequate" & Equipment == "Available"), n = 1e6)
```

### Compare Interventions

The `mutilated()` function implements the do-operator -- it forces a variable to a specific value and severs its incoming arrows. This lets us compare what happens when we *intervene* rather than just *observe*.

```{r}
#| label: intervention-comparison
#| echo: true

library(gRain)

# Status quo
p_baseline <- querygrain(as.grain(bn_tz), nodes = "NMR")$NMR["High"]

# do(Staffing = Adequate) -- what if we fix staffing everywhere?
bn_do_staff <- mutilated(bn_tz, evidence = list(Staffing = "Adequate"))
p_staff <- querygrain(as.grain(bn_do_staff), nodes = "NMR")$NMR["High"]

# do(Equipment = Available) -- what if we fix equipment everywhere?
bn_do_equip <- mutilated(bn_tz, evidence = list(Equipment = "Available"))
p_equip <- querygrain(as.grain(bn_do_equip), nodes = "NMR")$NMR["High"]

# Compare all three scenarios
data.frame(
  Scenario = c("Status Quo", "Fix Staffing", "Fix Equipment"),
  P_High_NMR = round(c(p_baseline, p_staff, p_equip), 3)
)
```

## Key Takeaways

::: {.key-takeaway}
**Bayesian network = DAG + CPTs.** The diagram says what causes what; the tables say how strongly. Together they answer any probability question you can ask.

**Natural frequencies first, formulas second.** "34 out of 40 facilities" is easier to grasp than P(Q|S) = 0.85. Always ground the math in real counts.

**Bayes' rule flips the direction.** You see the outcome (high mortality) and reason backward to the most likely cause (staffing? equipment? both?). The tree diagram makes this visual and intuitive.

**The do-operator reveals causal effects.** Observing that well-staffed facilities have lower mortality is not the same as knowing that *fixing* staffing will lower mortality. Graph surgery (the do-operator) isolates the true causal effect.
:::

## Looking Ahead

Next in **Chapter 5 -- Situational Analysis**, we put this Bayesian network to work on a real diagnostic puzzle in Ethiopia:

- Running the model *backward* -- from observed outcomes to hidden causes
- Updating beliefs as multiple pieces of evidence arrive, one at a time
- The **explaining away** phenomenon -- why confirming one cause makes a competing cause less likely
- A complete district diagnosis workflow using Bayesian inference
