---
title: "Chapter 4 — Putting Numbers on the Picture"
subtitle: "Joyce, the last 100 mothers, and lookup tables"
---

## The mystery

Tanzania's national maternal mortality ratio is **524 per 100,000 live births** *(WHO 2017)*. But Joyce — a midwife in Mwanza Region — does not deliver 100,000 births. She delivers about 100 a year. A number that's right at the country level can be useless at the bedside. We need a way to take the picture from Chapters 2–3 and turn each arrow into a number Joyce can actually use.

::: {.callout-note}
**The one thing to remember from this chapter:** A picture from Chapter 2 + a lookup table on every arrow = a Bayesian network. The picture tells you the structure; the counts tell you the strengths.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch04-quantitative-models/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch04-quantitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — build Joyce's first Bayesian network

In the lecture, we built the smallest possible picture-with-numbers for Joyce's facility: **Anaemia → BadOutcome**, with the counts from her last 100 mothers (5 anaemia mothers had a bad outcome, 1 non-anaemia mother had one). In this workshop you'll do the same in R using the `bnlearn` package — and then ask the network the same forward and backward questions Joyce asks at the bedside.

### Setting up

```{r}
#| label: ch04-setup
# bnlearn handles Bayesian networks (pictures with lookup tables on the arrows).
# gRain lets us ask the network specific questions.
library(bnlearn)
library(gRain)
```

### Joyce's picture, with structure only

```{r}
#| label: ch04-structure
# First, just the picture (no numbers yet).
# Two boxes, one arrow — the smallest possible Bayesian network.
joyce_dag <- model2network("[Anaemia][BadOutcome|Anaemia]")

# Show the structure
plot(joyce_dag)
```

What you should see: two boxes with one arrow. Anaemia → BadOutcome. The same shape as the lecture diagram.

### Now add Joyce's lookup table

The counts come straight from Joyce's logbook — exactly the table from the lecture.

```{r}
#| label: ch04-tables
# Joyce had 40 anaemia mothers and 60 non-anaemia mothers out of 100.
# Think of these as "40 out of 100", not "0.40".
prob_anaemia <- array(
  c(0.40, 0.60),
  dim = 2,
  dimnames = list(Anaemia = c("yes", "no"))
)

# Of every 40 anaemia mothers, 5 had a bad outcome (5/40 = 12.5%) and 35 did not.
# Of every 60 non-anaemia mothers, 1 had a bad outcome (1/60 ≈ 1.7%) and 59 did not.
prob_bad_given_anaemia <- array(
  c(0.125, 0.875,    # Anaemia = yes  → bad 5/40 = 12.5%, good 87.5%
    0.0167, 0.9833), # Anaemia = no   → bad 1/60 ≈ 1.7%,  good 98.3%
  dim = c(2, 2),
  dimnames = list(
    BadOutcome = c("yes", "no"),
    Anaemia    = c("yes", "no")
  )
)

# Stitch the table onto the picture.
joyce_bn <- custom.fit(joyce_dag, dist = list(
  Anaemia    = prob_anaemia,
  BadOutcome = prob_bad_given_anaemia
))
```

### Ask the network Joyce's bedside questions

The whole point of the picture-with-numbers is that the *same* lookup table answers different questions when you ask it from different directions.

```{r}
#| label: ch04-query
# Convert to gRain format so we can run queries.
joyce_grain <- as.grain(joyce_bn)

# FORWARD question: a new mother walks in with anaemia.
# What's the chance of a bad outcome?
querygrain(setEvidence(joyce_grain,
                       evidence = list(Anaemia = "yes")),
           nodes = "BadOutcome")

# BACKWARD question: a bad outcome happened.
# What's the chance the mother had anaemia?
querygrain(setEvidence(joyce_grain,
                       evidence = list(BadOutcome = "yes")),
           nodes = "Anaemia")
```

What you should see:

- **Forward question** returns about **12.5%** — that's 5 out of 40 anaemia mothers having a bad outcome.
- **Backward question** returns about **83%** — that's 5 of the 6 bad outcomes happening in anaemia mothers.

Same table, two questions, two answers. That is the magic of putting numbers on the picture. The arrow can run in both directions when you ask the network a question.

## Try it yourself

In a Tanzanian district, 100 women were tested for HIV during antenatal care last month. 20 tested positive. Of those 20, 3 had a low-birth-weight baby. Of the 80 who tested negative, 6 had a low-birth-weight baby.

Build a tiny lookup table. Then answer: a new HIV-positive mother is how many times more likely to have a low-birth-weight baby than an HIV-negative mother?
