---
title: "Chapter 4 — Putting Numbers on the Picture"
subtitle: "Joyce, the last 100 mothers, and lookup tables"
---

## The mystery

Tanzania's national maternal mortality ratio is **524 per 100,000 live births** *(WHO 2017)*. But Joyce — a midwife in Mwanza Region — does not deliver 100,000 births. She delivers about 100 a year. A number that's right at the country level can be useless at the bedside. We need a way to take the picture from Chapters 2–3 and turn each arrow into a number Joyce can actually use.

::: {.callout-note}
**The one thing to remember from this chapter:** A picture from Chapter 2 + a lookup table on every arrow = a Bayesian network. The picture tells you the structure; the counts tell you the strengths.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch04-quantitative-models/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch04-quantitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — build Joyce's first Bayesian network

In the lecture, we built a small two-arrow picture for Joyce's facility: **Anaemia → Bleeding → BadOutcome**, and we filled in counts from her last 100 mothers. In this workshop you'll do the same in R using the `bnlearn` package — and then ask the network the same questions Joyce asks at the bedside.

### Setting up

```{r}
#| label: ch04-setup
# bnlearn handles Bayesian networks (pictures with lookup tables on the arrows).
# gRain lets us ask the network specific questions.
library(bnlearn)
library(gRain)
```

### Joyce's picture, with structure only

```{r}
#| label: ch04-structure
# First, just the picture (no numbers yet).
# This is the same kind of structure we drew in Chapter 2.
joyce_dag <- model2network("[Anaemia][Bleeding|Anaemia][BadOutcome|Bleeding]")

# Show the structure
plot(joyce_dag)
```

What you should see: three boxes in a chain. Anaemia → Bleeding → BadOutcome. Same shape as the lecture diagram.

### Now add Joyce's lookup tables

Here are the counts straight from Joyce's logbook, in the same form as the lecture table:

```{r}
#| label: ch04-tables
# Joyce had 40 anaemia mothers and 60 non-anaemia mothers out of 100.
# That's 0.40 vs 0.60 — but think of it as "40 out of 100".
prob_anaemia <- array(
  c(0.40, 0.60),
  dim = 2,
  dimnames = list(Anaemia = c("yes", "no"))
)

# Of every 40 anaemia mothers, 24 had bleeding (60%); of every 60 non-anaemia
# mothers, only 12 had bleeding (20%).
prob_bleeding_given_anaemia <- array(
  c(0.60, 0.40,   # Anaemia = yes  → bleeding 60%, no bleeding 40%
    0.20, 0.80),  # Anaemia = no   → bleeding 20%, no bleeding 80%
  dim = c(2, 2),
  dimnames = list(
    Bleeding = c("yes", "no"),
    Anaemia  = c("yes", "no")
  )
)

# Of every 36 women who bled, 5 had a bad outcome (~14%); of every 64 who
# didn't bleed, 1 had a bad outcome (~1.5%).
prob_bad_given_bleeding <- array(
  c(0.14, 0.86,
    0.015, 0.985),
  dim = c(2, 2),
  dimnames = list(
    BadOutcome = c("yes", "no"),
    Bleeding   = c("yes", "no")
  )
)

# Stitch the tables onto the picture.
joyce_bn <- custom.fit(joyce_dag, dist = list(
  Anaemia    = prob_anaemia,
  Bleeding   = prob_bleeding_given_anaemia,
  BadOutcome = prob_bad_given_bleeding
))
```

### Ask the network Joyce's bedside questions

```{r}
#| label: ch04-query
# Convert to gRain format so we can run queries.
joyce_grain <- as.grain(joyce_bn)

# Question 1: a new mother walks in with anaemia. What's the chance of a bad outcome?
querygrain(setEvidence(joyce_grain,
                       evidence = list(Anaemia = "yes")),
           nodes = "BadOutcome")

# Question 2 (the reverse): a mother had a bad outcome. What's the chance she had anaemia?
querygrain(setEvidence(joyce_grain,
                       evidence = list(BadOutcome = "yes")),
           nodes = "Anaemia")
```

What you should see: question 1 gives you a probability that's much higher than the 6% baseline — because the anaemia evidence has flowed *forward* through the picture. Question 2 gives you a probability that's much higher than the 40% baseline — because the bad-outcome evidence has flowed *backward* through the picture.

That's the magic of putting numbers on the picture. The arrows can run in both directions when you ask the network a question.

## Try it yourself

In a Tanzanian district, 100 women were tested for HIV during antenatal care last month. 20 tested positive. Of those 20, 3 had a low-birth-weight baby. Of the 80 who tested negative, 6 had a low-birth-weight baby.

Build a tiny lookup table. Then answer: a new HIV-positive mother is how many times more likely to have a low-birth-weight baby than an HIV-negative mother?
