---
title: "Chapter 5: Situational Analysis"
subtitle: "Running the model backward -- from what you observe to what is probably true"
---

## Learning Objectives

::: {.learning-objectives}
- Compute **marginal probabilities** by summing across all possible parent combinations
- **Update beliefs with evidence** using Bayes' rule -- start with a prior, see evidence, get a posterior
- Trace how information flows through **serial chains, forks, and colliders** and explain why each behaves differently
- Apply the **explaining away** phenomenon -- when two causes compete to explain the same effect
- Run a complete **district diagnosis** in R, ranking the most likely upstream problems given observed outcomes
:::

## Key Concepts

::: {.key-concept}
#### Belief Updating

Chapter 4 ran models forward: "If staffing is low, what happens to mortality?" Situational analysis runs them backward: "Mortality is high -- what is probably causing it?" This is how doctors diagnose patients, how mechanics troubleshoot engines, and how health program managers in Ethiopia figure out why a district is struggling. You start with a prior belief (your best guess before seeing any data), observe evidence (the outcome data from the district), and compute a posterior belief (your updated guess). Bayes' rule is the engine that drives this update.
:::

::: {.key-concept .green}
#### Marginal Probabilities

A marginal probability answers: "What is the overall chance of this outcome, before I know anything specific?" To compute it, you sum over all possible parent states, weighting each by how likely that parent state is. For example, P(NMR = High) in Ethiopia averages across districts with good staffing, bad staffing, good equipment, bad equipment -- everything. The marginal is your baseline. It is where you start before any district-specific evidence arrives. Once evidence comes in, the marginal gets replaced by a posterior -- and the gap between the two tells you how much you learned.
:::

::: {.key-concept .orange}
#### Evidence Flow in the Three Triplet Structures

The three building blocks of every DAG each handle information differently. In a **serial chain** (Staffing -> Quality -> NMR), information flows from end to end -- but if you already know Quality, learning about Staffing tells you nothing new about NMR. The chain is blocked. In a **fork** (Quality <- Disease Burden -> Mortality), the common cause creates a spurious link -- but conditioning on Disease Burden removes it. In a **collider** (Staffing -> Quality <- Equipment), the two causes are independent by default. But the moment you observe Quality, they become linked. These three rules scale to any size network.
:::

::: {.key-concept .red}
#### Explaining Away

This is the most counterintuitive idea in the entire course. Picture a collider: Staffing -> Quality <- Equipment. Normally, staffing levels and equipment availability have nothing to do with each other. But suppose you learn that a district in Ethiopia has good-quality care despite low staffing. Suddenly, equipment availability must be high -- because *something* has to explain the good quality, and if it is not staffing, it must be equipment. Confirming one cause makes the competing cause less likely. Disconfirming one makes the other more likely. This "explaining away" effect only kicks in when you condition on the collider (the shared effect).
:::

::: {.key-concept .teal}
#### Diagnostic Inference

Diagnostic inference is the practical payoff of everything above. Given an observation -- say, a district in Ethiopia's Amhara region reporting high neonatal mortality -- you feed that evidence into the Bayesian network and read off the posterior probabilities for every upstream cause. Which shifted the most from its prior? That is your top suspect. This turns a vague question ("Why is this district struggling?") into a ranked list of probable causes, ordered by how much the evidence shifted each belief. It is data-driven diagnosis.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch05-situational-analysis/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch05-situational-analysis/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

We build a five-node Bayesian network for Ethiopia's health system, then use it to diagnose districts -- reasoning backward from observed mortality to the most likely upstream problems.

### Build the Ethiopia Network

Three root nodes (Staffing, Equipment, Training) feed into Quality, which feeds into NMR. The CPTs are calibrated to publicly available data from Ethiopia's DHS and WHO reports.

```{r}
#| label: build-network
#| echo: true

library(bnlearn)
library(gRain)

# Five-node DAG for Ethiopia
dag <- model2network(
  "[Staffing][Equipment][Training][Quality|Staffing:Equipment:Training][NMR|Quality]"
)

# Root node CPTs (based on Ethiopia DHS/WHO public data)
cpt_s <- array(c(0.40, 0.60), dim = 2,
  dimnames = list(Staffing = c("Adequate", "Low")))

cpt_e <- array(c(0.35, 0.65), dim = 2,
  dimnames = list(Equipment = c("Available", "Unavailable")))

cpt_t <- array(c(0.50, 0.50), dim = 2,
  dimnames = list(Training = c("Completed", "Not_Completed")))

# Quality CPT: 8 rows (2 x 2 x 2 parent combinations)
cpt_q <- array(
  c(0.95, 0.05,  0.80, 0.20,  0.65, 0.35,  0.45, 0.55,
    0.55, 0.45,  0.35, 0.65,  0.20, 0.80,  0.08, 0.92),
  dim = c(2, 2, 2, 2),
  dimnames = list(Quality = c("Good", "Poor"),
    Staffing = c("Adequate", "Low"),
    Equipment = c("Available", "Unavailable"),
    Training = c("Completed", "Not_Completed")))

# NMR CPT
cpt_nmr <- array(c(0.85, 0.15, 0.30, 0.70), dim = c(2, 2),
  dimnames = list(NMR = c("Low", "High"), Quality = c("Good", "Poor")))

# Assemble the fitted network
bn_eth <- custom.fit(dag, list(
  Staffing = cpt_s, Equipment = cpt_e, Training = cpt_t,
  Quality = cpt_q, NMR = cpt_nmr))
```

### Forward vs. Backward Queries

Forward: "Given the cause, predict the effect." Backward: "Given the effect, infer the cause." Backward queries are the heart of situational analysis.

```{r}
#| label: forward-backward
#| echo: true

# FORWARD: P(NMR = High | Staffing = Low)
# "If staffing is low, how likely is high mortality?"
cpquery(bn_eth,
  event = (NMR == "High"), evidence = (Staffing == "Low"), n = 1e6)

# BACKWARD: P(Staffing = Low | NMR = High)
# "Mortality is high -- how likely is it that staffing is the problem?"
cpquery(bn_eth,
  event = (Staffing == "Low"), evidence = (NMR == "High"), n = 1e6)

# Marginal baseline (no evidence at all)
cpquery(bn_eth, event = (NMR == "High"), evidence = TRUE, n = 1e6)
```

### Compare Prior vs. Posterior

When we observe high NMR in a district, how much does each root cause shift from its baseline? The biggest shift points to the most likely culprit.

```{r}
#| label: prior-posterior
#| echo: true

library(ggplot2)

# Compute posteriors for each root node given NMR = High
priors    <- c(Staffing = 0.60, Equipment = 0.65, Training = 0.50)
posteriors <- c(
  cpquery(bn_eth, (Staffing == "Low"), (NMR == "High"), n = 1e6),
  cpquery(bn_eth, (Equipment == "Unavailable"), (NMR == "High"), n = 1e6),
  cpquery(bn_eth, (Training == "Not_Completed"), (NMR == "High"), n = 1e6))

# Side-by-side comparison
results <- data.frame(
  Variable = rep(c("Staffing\n(Low)", "Equipment\n(Unavail)", "Training\n(Not Done)"), 2),
  Belief   = rep(c("Prior", "Posterior"), each = 3),
  Prob     = c(priors, posteriors))

results$Belief <- factor(results$Belief, levels = c("Prior", "Posterior"))

ggplot(results, aes(x = Variable, y = Prob, fill = Belief)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Prior" = "#00A1DF", "Posterior" = "#E24A3F")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Belief Updating: Prior vs. Posterior Given NMR = High",
       subtitle = "All 'bad' states become more likely after observing high mortality",
       x = NULL, y = "Probability") +
  theme_minimal(base_size = 14)
```

### District Diagnosis Function

This reusable function takes a fitted network and a set of observations, then ranks every root node by how much its belief shifted. The biggest movers are your top diagnostic priorities.

```{r}
#| label: diagnosis-function
#| echo: true

diagnose_district <- function(bn, evidence_expr, n_samples = 1e6) {
  nodes_info <- data.frame(
    Node = c("Staffing", "Equipment", "Training"),
    Bad_State = c("Low", "Unavailable", "Not_Completed"),
    Prior = c(0.60, 0.65, 0.50))

  nodes_info$Posterior <- c(
    cpquery(bn, (Staffing == "Low"), evidence_expr, n = n_samples),
    cpquery(bn, (Equipment == "Unavailable"), evidence_expr, n = n_samples),
    cpquery(bn, (Training == "Not_Completed"), evidence_expr, n = n_samples))

  nodes_info$Shift <- nodes_info$Posterior - nodes_info$Prior
  nodes_info[order(-abs(nodes_info$Shift)), ]
}

# Diagnose a district in Amhara region with high NMR
diagnose_district(bn_eth, (NMR == "High"))
```

## Key Takeaways

::: {.key-takeaway}
**Situational analysis runs the model backward.** Instead of predicting outcomes from causes, you observe outcomes and infer the most likely causes. This is how real diagnosis works.

**Marginal probabilities are your baseline.** The gap between the marginal (before evidence) and the posterior (after evidence) tells you exactly how much you learned from each observation.

**The three triplet structures control information flow.** Chains and forks are blocked by conditioning on the middle node. Colliders are opened by it. Get this wrong and you introduce bias instead of removing it.

**Explaining away is real and powerful.** When two independent causes share a common effect, observing the effect makes them compete. This is not a bug -- it is the correct way to reason under uncertainty.
:::

## Looking Ahead

Next in **Chapter 6 -- Simpson's Paradox**, we tackle the most dangerous trap in data analysis:

- When aggregated data tells you the *opposite* of the truth
- Why you must always condition on confounders but never on mediators
- The Prosecutor's Fallacy -- confusing P(evidence | hypothesis) with P(hypothesis | evidence)
- A cross-country comparison across Ethiopia, Rwanda, Kenya, and Tanzania that gets the wrong answer without causal reasoning
