---
title: "Chapter 5 — When New Evidence Arrives"
subtitle: "Hiwot, three explanations, and the redistribution rule"
---

## The mystery

A regional health director in Amhara opens her email Monday morning. Maternal deaths in one of her districts jumped from **9 to 14** in a single quarter. Three explanations sit on the table within the hour: personnel attrition, a measles surge, or a reporting pipeline cleanup. Over the next two weeks, evidence will arrive piece by piece. Her belief has to update each time — without overreacting.

::: {.callout-note}
**The one thing to remember from this chapter:** Bayesian updating is just redistributing your "out of 100" between explanations — keeping the total at 100 — based on which ones the new evidence fits.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch05-situational-analysis/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch05-situational-analysis/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — update Hiwot's belief as evidence arrives

In the lecture, we walked through Hiwot's three competing explanations, started with a 33-33-33 prior, and watched the numbers move twice — once on Tuesday, once on Wednesday. In this workshop you'll do the same updating in R, using the same Bayesian network machinery from Chapter 4.

### Setting up

```{r}
#| label: ch05-setup
library(bnlearn)
library(gRain)
```

### A picture for the three competing explanations

```{r}
#| label: ch05-structure
# Three boxes: each is one of Hiwot's three explanations.
# All three feed into the observed rise in deaths.
hiwot_dag <- model2network("[Personnel][Measles][Reporting][DeathsRose|Personnel:Measles:Reporting]")

plot(hiwot_dag)
```

What you should see: three parents (Personnel, Measles, Reporting) and one child (DeathsRose). Each parent could on its own explain the rise — that's the *collider* shape from Chapter 2.

### Fill in the lookup tables

```{r}
#| label: ch05-tables
# Hiwot's prior: roughly 33-33-33 across the three explanations.
prior_personnel <- c(yes = 0.33, no = 0.67)
prior_measles   <- c(yes = 0.33, no = 0.67)
prior_reporting <- c(yes = 0.33, no = 0.67)

# DeathsRose is "yes" if any of the three parents is "yes" (a noisy-OR).
# IMPORTANT: R fills array() in column-major order — the FIRST dimension
# (DeathsRose) varies fastest, then Personnel, then Measles, then Reporting.
# Each pair below is (P(DeathsRose=yes), P(DeathsRose=no)) for one combination
# of the three parents.
deaths_table <- array(
  c(
    # Reporting = no, Measles = no
    0.05, 0.95,   # Personnel = no   (none of the three: baseline)
    0.70, 0.30,   # Personnel = yes  (personnel only)
    # Reporting = no, Measles = yes
    0.70, 0.30,   # Personnel = no   (measles only)
    0.95, 0.05,   # Personnel = yes  (personnel + measles)
    # Reporting = yes, Measles = no
    0.70, 0.30,   # Personnel = no   (reporting only)
    0.95, 0.05,   # Personnel = yes  (personnel + reporting)
    # Reporting = yes, Measles = yes
    0.95, 0.05,   # Personnel = no   (measles + reporting)
    0.99, 0.01    # Personnel = yes  (all three)
  ),
  dim = c(2, 2, 2, 2),
  dimnames = list(
    DeathsRose = c("yes", "no"),
    Personnel  = c("no", "yes"),
    Measles    = c("no", "yes"),
    Reporting  = c("no", "yes")
  )
)

hiwot_bn <- custom.fit(hiwot_dag, dist = list(
  Personnel  = prior_personnel,
  Measles    = prior_measles,
  Reporting  = prior_reporting,
  DeathsRose = deaths_table
))

hiwot_grain <- as.grain(hiwot_bn)
```

### Monday: Hiwot only knows the deaths rose

```{r}
#| label: ch05-monday
# Just the observation that deaths rose. No other evidence yet.
querygrain(setEvidence(hiwot_grain, evidence = list(DeathsRose = "yes")),
           nodes = c("Personnel", "Measles", "Reporting"))
```

You'll see all three explanations sitting roughly close to one another — the prior plus the simple fact "deaths rose".

### Tuesday: the chief of staff calls

The personnel story gains support; the measles story loses support.

```{r}
#| label: ch05-tuesday
querygrain(setEvidence(hiwot_grain,
                       evidence = list(DeathsRose = "yes",
                                       Measles    = "no")),
           nodes = c("Personnel", "Measles", "Reporting"))
```

Personnel is now meaningfully more likely than reporting.

### Wednesday: the HMIS officer confirms a coding cleanup

This is the *explaining away* moment from the lecture.

```{r}
#| label: ch05-wednesday
querygrain(setEvidence(hiwot_grain,
                       evidence = list(DeathsRose = "yes",
                                       Measles    = "no",
                                       Reporting  = "yes")),
           nodes = c("Personnel", "Measles", "Reporting"))
```

Watch what happens to **Personnel**. Even though no new evidence about personnel arrived, its probability *drops* — because reporting is now absorbing more of the rise. That's the redistribution rule from the lecture, doing the work without you.

## Try it yourself

A regional director in Kenya sees antenatal coverage in one county jump from 65% to 78% in a single quarter. Three explanations: a new SMS reminder programme, a DHIS2 coding change, or a measles vaccination drive that brought women into clinics where they were also screened for ANC. Start with 33-33-33. A phone call confirms the DHIS2 coding change is real. How do your numbers move? What's the next piece of evidence you should ask for?
