---
title: "Chapter 6 — When the Data Lies in Two Directions"
subtitle: "Hassan, Naomi, and Simpson's Paradox"
---

## The mystery

A Kenyan voucher programme looks like a 10-percentage-point success in the aggregate. Split by access to a clinic, the relationship *reverses* — the voucher does worse than no voucher in *every* access group. Same data. Opposite story. Which version should you trust?

::: {.callout-note}
**The one thing to remember from this chapter:** When data tells you opposite things at the aggregate and split levels, the data is not lying. The picture is what tells you which view to trust. A fork → split. A mediator → don't.
:::

## Main lecture

::: {.slide-embed}
<iframe src="../ch06-business-financials/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch06-business-financials/index.html" target="_blank">Open lecture slides full-screen &rarr;</a>
</div>
:::

## Tutorial / breakout session

The tutorial deck takes the same ideas and retells them as a story — built for a live breakout session, not a lecture.

::: {.slide-embed}
<iframe src="../ch06-simpsons-paradox-tutorial/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch06-simpsons-paradox-tutorial/index.html" target="_blank">Open tutorial slides full-screen &rarr;</a>
</div>
:::

## R workshop — see Simpson's Paradox happen, then resolve it

In the lecture, we walked Hassan through the trap and back out using a picture. In this workshop you'll build the exact same dataset in R, watch the paradox appear in your console, and then watch how the choice of picture determines whether you should disaggregate.

### Setting up

```{r}
#| label: ch06-setup
library(dplyr)
library(dagitty)
library(ggdag)
```

### Build the dataset from the lecture

The numbers are exactly those from the lecture table — 100 voucher households and 100 non-voucher households, split by whether the household lives in an easy-to-reach area (close to a registered facility) or a hard-to-reach area.

```{r}
#| label: ch06-data
voucher_data <- tibble::tribble(
  ~Access,         ~Voucher, ~ClinicDelivery, ~n,
  "easy",          "yes",    "yes",           60,
  "easy",          "yes",    "no",            10,
  "easy",          "no",     "yes",           45,
  "easy",          "no",     "no",            5,
  "hard",          "yes",    "yes",           5,
  "hard",          "yes",    "no",            25,
  "hard",          "no",     "yes",           10,
  "hard",          "no",     "no",            40
) |>
  tidyr::uncount(n)

nrow(voucher_data)  # 200 rows
```

### The aggregate view (what Hassan saw first)

```{r}
#| label: ch06-aggregate
voucher_data |>
  group_by(Voucher) |>
  summarise(
    pct_clinic = mean(ClinicDelivery == "yes")
  )
```

You'll see voucher = 65%, no voucher = 55%. A clean +10 percentage points in favour of the voucher.

### The split view (what Naomi told Hassan to look at)

```{r}
#| label: ch06-split
voucher_data |>
  group_by(Access, Voucher) |>
  summarise(
    pct_clinic = mean(ClinicDelivery == "yes"),
    .groups = "drop"
  )
```

Now look at every Access × Voucher cell:

- **Easy to reach**: voucher 86%, no voucher 90% — voucher slightly *worse*
- **Hard to reach**: voucher 17%, no voucher 20% — voucher slightly *worse*

**In both access groups, voucher does worse than no voucher.** Same data. Opposite story. Simpson's Paradox is staring at you from your console.

### Let the picture decide which view to trust

```{r}
#| label: ch06-picture
# Hassan's picture: access is upstream of voucher AND of clinic delivery.
# (The voucher rolled out where existing clinics could redeem it, so the
# programme is concentrated in easier-to-reach areas — and easier-to-reach
# areas have higher clinic delivery for many other reasons too.)
# That makes access a confounder — exactly the fork shape from Chapter 2.
voucher_dag <- dagitty('dag {
  Access  -> Voucher
  Access  -> ClinicDelivery
  Voucher -> ClinicDelivery
}')

ggdag(voucher_dag, text_size = 4) + theme_dag()

# Ask the picture: to estimate Voucher -> ClinicDelivery, what do we need to adjust for?
adjustmentSets(voucher_dag,
               exposure = "Voucher",
               outcome  = "ClinicDelivery")
```

R will tell you: **Access**. The picture says access is a confounder, so you have to adjust for it — meaning the *split* view is the right one. The aggregate is misleading.

### What if the picture were different?

```{r}
#| label: ch06-mediator
# Alternative picture: the voucher itself creates effective access (e.g., it
# includes a transport subsidy that pays for the motorbike). In this picture
# Access is downstream of Voucher — a mediator. Adjusting for it would BLOCK
# the very effect you're trying to measure.
mediator_dag <- dagitty('dag {
  Voucher -> Access
  Access  -> ClinicDelivery
  Voucher -> ClinicDelivery
}')

adjustmentSets(mediator_dag,
               exposure = "Voucher",
               outcome  = "ClinicDelivery")
```

R will return an empty set — meaning, in *this* picture, you should *not* adjust for access. The aggregate is right, the split is wrong.

**Same data. Two pictures. Opposite advice.** This is exactly the point from the lecture: the data alone cannot pick a winner. The picture has to.

## Try it yourself

A national report says women who use mobile money are more likely to seek antenatal care. Then someone splits the same data by region and finds that within every region, mobile money users are *less* likely to seek antenatal care. Before you decide which version to publish: draw the picture. Where does region go? Is region a confounder, or a mediator? What is the one piece of evidence that would settle it?
