---
title: "Chapter 6 — When the Data Lies in Two Directions"
subtitle: "Hassan, Naomi, and Simpson's Paradox"
---

## The mystery

A Kenyan voucher programme looks like a 26-percentage-point success in the aggregate. Split by income, the relationship *reverses* — the voucher does worse than no voucher in *every* income group. Same data. Opposite story. Which version should you trust?

::: {.callout-note}
**The one thing to remember from this chapter:** When data tells you opposite things at the aggregate and split levels, the data is not lying. The picture is what tells you which view to trust. A fork → split. A mediator → don't.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch06-simpsons-paradox/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch06-simpsons-paradox/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — see Simpson's Paradox happen, then resolve it

In the lecture, we walked Hassan through the trap and back out using a picture. In this workshop you'll see the paradox appear in real numbers and watch how the choice of adjustment variable flips the answer.

### Setting up

```{r}
#| label: ch06-setup
library(dplyr)
library(dagitty)
library(ggdag)
```

### Build the dataset from the lecture

```{r}
#| label: ch06-data
# Same numbers as the lecture table.
voucher_data <- tibble::tribble(
  ~Income,    ~Voucher,    ~ClinicDelivery, ~n,
  "low",      "yes",       "yes",           40,
  "low",      "yes",       "no",            20,
  "low",      "no",        "yes",           35,
  "low",      "no",        "no",            15,
  "high",     "yes",       "yes",           38,
  "high",     "yes",       "no",            2,
  "high",     "no",        "yes",           17,
  "high",     "no",        "no",            33
) |>
  tidyr::uncount(n)

nrow(voucher_data)  # 200 rows
```

### The aggregate view (what Hassan saw first)

```{r}
#| label: ch06-aggregate
voucher_data |>
  group_by(Voucher) |>
  summarise(
    pct_clinic = mean(ClinicDelivery == "yes")
  )
```

You'll see voucher = 78%, no voucher = 52%. A clean +26 percentage points.

### The split view (what Naomi told Hassan to look at)

```{r}
#| label: ch06-split
voucher_data |>
  group_by(Income, Voucher) |>
  summarise(
    pct_clinic = mean(ClinicDelivery == "yes"),
    .groups = "drop"
  )
```

Now look at every Income × Voucher cell. **In both income groups, voucher does worse than no voucher.** Same data. Opposite story. Simpson's Paradox is staring at you from your console.

### Let the picture decide which view to trust

```{r}
#| label: ch06-picture
# Hassan's picture: income is upstream of voucher AND of clinic delivery.
# That makes income a confounder — exactly the fork shape from Chapter 2.
voucher_dag <- dagitty('dag {
  Income  -> Voucher
  Income  -> ClinicDelivery
  Voucher -> ClinicDelivery
}')

ggdag(voucher_dag, text_size = 4) + theme_dag()

# Ask the picture: to estimate Voucher -> ClinicDelivery, what do we need to adjust for?
adjustmentSets(voucher_dag,
               exposure = "Voucher",
               outcome  = "ClinicDelivery")
```

R will tell you: **Income**. The picture says income is a confounder, so you have to adjust for it — meaning the *split* view is the right one. The aggregate is misleading.

### What if the picture were different?

```{r}
#| label: ch06-mediator
# Alternative picture: voucher causes income (e.g., the voucher includes a cash transfer).
# In this picture, income is a mediator — adjusting for it BLOCKS the very effect you're measuring.
mediator_dag <- dagitty('dag {
  Voucher -> Income
  Income  -> ClinicDelivery
  Voucher -> ClinicDelivery
}')

adjustmentSets(mediator_dag,
               exposure = "Voucher",
               outcome  = "ClinicDelivery")
```

R will return an empty set — meaning, in *this* picture, you should *not* adjust for income. The aggregate is right, the split is wrong.

**Same data. Two pictures. Opposite advice.** This is exactly the point from the lecture: the data alone cannot pick a winner. The picture has to.

## Try it yourself

A national report says women who use mobile money are more likely to seek antenatal care. Then someone splits the same data by region and finds that within every region, mobile money users are *less* likely to seek antenatal care. Before you decide which version to publish: draw the picture. Where does region go? Is region a confounder, or a mediator? What is the one piece of evidence that would settle it?
