---
title: "Chapter 6: Simpson's Paradox"
subtitle: "When aggregated data lies -- and how causal reasoning sets the record straight"
---

## Learning Objectives

::: {.learning-objectives}
- Identify **Simpson's Paradox** -- when a trend in aggregated data reverses once you split by a confounding variable
- Distinguish a **confounder** (condition on it) from a **mediator** (leave it alone) and explain why getting this wrong flips your conclusions
- Spot the **Prosecutor's Fallacy** -- confusing "how likely is the evidence if the hypothesis is true" with "how likely is the hypothesis given the evidence"
- Conduct a **cross-country comparison** across Ethiopia, Rwanda, Kenya, and Tanzania that avoids the paradox
- Use R to simulate, visualize, and correctly analyze data that exhibits these traps
:::

## Key Concepts

::: {.key-concept}
#### Simpson's Paradox

Imagine comparing a health intervention across Ethiopia, Rwanda, Kenya, and Tanzania. In every single country, facilities with the intervention have better outcomes. But when you pool all the data together, the intervention appears to *hurt*. How? Because the intervention was deployed mostly in high-burden countries with worse baseline outcomes. The aggregate mixes apples and oranges. Simpson's Paradox is not rare or exotic -- it lurks in any cross-country or cross-site comparison where a hidden variable drives both the treatment assignment and the outcome. The fix is not more data. It is causal reasoning.
:::

::: {.key-concept .green}
#### Confounder vs. Mediator

A **confounder** is a common cause of both the treatment and the outcome -- like disease burden, which determines both where interventions are deployed and how high mortality is. You must condition on confounders to remove the spurious association they create. A **mediator** sits on the causal path between treatment and outcome -- like quality of care, which is *how* the intervention works. You must *not* condition on mediators, because doing so blocks the very effect you are trying to measure. In DAG terms: condition on the fork's center node, leave the chain's middle node alone. Mix them up and you get Simpson's Paradox -- or you erase a real treatment effect.
:::

::: {.key-concept .orange}
#### The Prosecutor's Fallacy

A screening test for birth complications has 90% sensitivity -- it catches 90 out of 100 true cases. A prosecutor might claim: "The test was positive, so there is a 90% chance the complication is present." Wrong. That confuses P(positive test | complication) with P(complication | positive test). If the complication is rare (say 5% prevalence), most positive tests are false alarms. Bayes' rule shows that P(complication | positive test) is only about 24%, not 90%. The base rate matters enormously. This fallacy appears whenever people mistake the likelihood for the posterior -- in courtrooms, in medical screening, and in program evaluation.
:::

::: {.key-concept .red}
#### Cross-Country Comparison

Comparing health outcomes across Ethiopia, Rwanda, Kenya, and Tanzania is essential for allocating resources wisely -- but naive comparisons are dangerous. Each country has a different baseline burden, different health system capacity, and different population structure. A country with low cost per outcome might simply have had less severe problems to begin with. Adjusting for confounders -- baseline mortality, urbanization, health worker density -- can completely reverse which country looks most "efficient." The DAG tells you which variables to adjust for and which to leave alone.
:::

::: {.key-concept .teal}
#### Causal Adjustment Fixes the Paradox

Once you draw the DAG, the solution is mechanical. Identify every backdoor path between treatment and outcome. Find the confounders along those paths. Condition on them (stratify, or include them in a regression). The adjusted estimate removes the bias and reveals the true causal effect. In the cross-country case, adjusting for baseline burden shows that the intervention helps in every country -- exactly what the within-country data already showed. The paradox disappears because you are no longer mixing high-burden and low-burden contexts.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch06-business-financials/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch06-business-financials/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

We simulate a dataset where Simpson's Paradox appears, visualize it, and show how stratifying by the confounder (or using a regression that includes it) recovers the true treatment effect.

### Simulate the Paradox

We create data for four countries. In every country, the intervention improves survival. But because high-burden countries receive more of the intervention, the pooled data reverses the effect.

```{r}
#| label: simulate-paradox
#| echo: true

library(tidyverse)
set.seed(42)

n <- 2000

sim_data <- tibble(
  # Confounder: country burden group
  burden = sample(c("High-burden", "Low-burden"), n, replace = TRUE,
                  prob = c(0.6, 0.4)),
  # Treatment is more common in high-burden countries (the confounder at work)
  treated = case_when(
    burden == "High-burden" ~ rbinom(n, 1, 0.75),
    burden == "Low-burden"  ~ rbinom(n, 1, 0.25)),
  # Outcome depends on BOTH burden and treatment
  # Treatment always helps (+7pp), but high-burden baseline is worse
  surv_prob = case_when(
    burden == "High-burden" & treated == 1 ~ 0.68,
    burden == "High-burden" & treated == 0 ~ 0.61,
    burden == "Low-burden"  & treated == 1 ~ 0.85,
    burden == "Low-burden"  & treated == 0 ~ 0.78),
  survived = rbinom(n, 1, surv_prob))

# Pooled result (misleading -- treatment looks harmful)
sim_data |> group_by(treated) |>
  summarise(survival = mean(survived), .groups = "drop")

# Stratified result (correct -- treatment helps in both groups)
sim_data |> group_by(burden, treated) |>
  summarise(survival = mean(survived), .groups = "drop")
```

### Visualize Aggregated vs. Stratified

Side-by-side bar charts make the paradox unmistakable. Same data, opposite conclusions.

```{r}
#| label: viz-paradox
#| echo: true
#| fig-width: 10
#| fig-height: 5

library(patchwork)

sim_data <- sim_data |>
  mutate(group = ifelse(treated == 1, "Treated", "Control"))

# Pooled view (misleading)
p1 <- ggplot(sim_data, aes(x = group, y = survived, fill = group)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6) +
  scale_fill_manual(values = c("Treated" = "#005CB9", "Control" = "#FA7650")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Pooled: Treatment Looks Harmful", y = "Survival Rate", x = NULL) +
  theme_minimal(base_size = 14) + theme(legend.position = "none")

# Stratified view (correct)
p2 <- ggplot(sim_data, aes(x = group, y = survived, fill = group)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6) +
  facet_wrap(~ burden) +
  scale_fill_manual(values = c("Treated" = "#005CB9", "Control" = "#FA7650")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Stratified: Treatment Helps Both Groups", y = "Survival Rate", x = NULL) +
  theme_minimal(base_size = 14) + theme(legend.position = "none")

p1 + p2 + plot_annotation(
  title = "Simpson's Paradox: Same Data, Opposite Conclusions")
```

### Regression: Naive vs. Adjusted

The naive regression omits the confounder and gets the sign wrong. The adjusted regression includes it and recovers the true positive effect.

```{r}
#| label: regression-fix
#| echo: true

# Naive: treatment coefficient is NEGATIVE (wrong!)
naive <- lm(survived ~ treated, data = sim_data)

# Adjusted: include the confounder, coefficient is POSITIVE (correct)
adjusted <- lm(survived ~ treated + burden, data = sim_data)

# Compare side by side
tibble(
  Model = c("Naive (no confounder)", "Adjusted (with confounder)"),
  Treatment_Effect = c(coef(naive)["treated"], coef(adjusted)["treated"]),
  Verdict = c("Treatment looks harmful", "Treatment is beneficial"))
```

### Prosecutor's Fallacy Calculator

This function shows how base rate (prevalence) changes the meaning of a positive test result. High sensitivity does not mean high predictive value when the condition is rare.

```{r}
#| label: bayes-calculator
#| echo: true

bayes_ppv <- function(sensitivity, specificity, prevalence) {
  p_pos <- (sensitivity * prevalence) + ((1 - specificity) * (1 - prevalence))
  ppv   <- (sensitivity * prevalence) / p_pos
  tibble(
    Sensitivity = scales::percent(sensitivity),
    Specificity = scales::percent(specificity),
    Prevalence  = scales::percent(prevalence),
    PPV         = scales::percent(ppv, accuracy = 0.1),
    Fallacy_Claims = scales::percent(sensitivity))
}

# Rare condition (5% prevalence) -- PPV is only ~24%
bayes_ppv(0.90, 0.85, 0.05)

# Common condition (20% prevalence) -- PPV rises to ~60%
bayes_ppv(0.90, 0.85, 0.20)
```

## Key Takeaways

::: {.key-takeaway}
**Simpson's Paradox is not a curiosity -- it is a constant threat.** Any time you pool data across groups that differ in baseline risk, the aggregate can reverse the true effect. Draw the DAG first.

**Confounders vs. mediators: the most important distinction in causal analysis.** Condition on confounders (common causes). Never condition on mediators (the causal pathway). Mixing them up produces exactly wrong conclusions.

**The Prosecutor's Fallacy kills intuition.** A 90% sensitive test does not mean a 90% chance the condition is present. Base rates matter. Always use Bayes' rule.

**Cross-country comparisons require causal adjustment.** Naive rankings by cost or outcomes will send resources to the wrong places. Adjusting for baseline burden and health system capacity can flip the rankings entirely.
:::

## Looking Ahead

Next in **Chapter 7 -- Decision Analysis**, we move from inference to action in Kenya:

- **Influence diagrams** -- adding decisions and objectives to causal models
- **Expected value** -- choosing the intervention that maximizes lives saved under uncertainty
- **Watching vs. doing** -- why the do-operator matters for real-world decisions
- **Value of information** -- how much should you pay for a pilot study before committing resources?
