---
title: "Chapter 7: Decision Analysis"
subtitle: "Influence diagrams, expected value, and knowing when better information is worth the wait"
---

## Learning Objectives

::: {.learning-objectives}
- Construct a **decision map (influence diagram)** with decision nodes, chance nodes, and objective nodes
- Compute the **expected value** of competing intervention strategies under uncertainty
- Explain the difference between **watching and doing** -- why observing a correlation is not the same as causing a change
- Apply **graph surgery (cutting the arrows)** to isolate the true causal effect of an intervention
- Calculate the **Expected Value of Perfect Information (EVPI)** to decide whether a pilot study or survey is worth funding
:::

## Key Concepts

::: {.key-concept}
#### Decision Maps (Influence Diagrams)

Up to now, our causal models have contained only chance nodes -- variables whose values are uncertain. A decision map adds two new ingredients. **Decision nodes** (drawn as rectangles) represent choices you control: which intervention to fund, where to deploy staff, whether to run a pilot first. **Objective nodes** (drawn as hexagons) represent the outcomes you care about: lives saved, cost-effectiveness, equity. Arrows into the objective node show what drives it. Arrows into a decision node show what information is available when the choice is made. In Kenya, a decision map might show a health ministry choosing between workforce investment, equipment procurement, or a split strategy -- with the objective of maximizing neonatal lives saved across 47 counties.
:::

::: {.key-concept .green}
#### Expected Value

When outcomes are uncertain, you cannot guarantee the best result. But you can choose the option with the highest *average* payoff. Expected value (EV) multiplies each possible outcome by its probability and adds them up. If a workforce strategy in Kenya saves 3,600 lives when government support is high (probability 0.40) and 1,000 lives when it is low (probability 0.60), the EV is 3,600 x 0.40 + 1,000 x 0.60 = 2,040 lives. Compare that to every alternative and pick the highest EV. It does not guarantee the best outcome every time, but across many decisions it is the smartest long-run strategy.
:::

::: {.key-concept .orange}
#### Watching vs. Doing

This is one of the biggest ideas in the course. **Watching** means observing a pattern in existing data: "Facilities with more equipment have lower mortality." **Doing** means intervening: "If we *give* facilities more equipment, will mortality drop?" These are different questions with potentially different answers. Why? Because what you observe might be confounded. Well-resourced counties may have both more equipment *and* better staffing, so the equipment-mortality correlation partly reflects staffing. The do-operator strips away the confounding by asking: "If we force equipment to be available regardless of everything else, what happens to mortality?" Only the do-operator gives you the causal effect.
:::

::: {.key-concept .red}
#### Cutting the Arrows (Graph Surgery)

Graph surgery is how we implement "doing" in a Bayesian network. To compute the effect of do(Equipment = Available), you delete all arrows pointing *into* the Equipment node and set it to "Available." This severs the link between Equipment and its causes (like county budget or supply chain capacity), isolating the pure causal effect of equipment on downstream outcomes. In the original network, P(NMR | Equipment = Available) includes confounding. In the surgically modified network, P(NMR | do(Equipment = Available)) is the true causal effect. Same network, different question, potentially different answer.
:::

::: {.key-concept .teal}
#### Value of Information

Before committing resources in Kenya, should you run a pilot project first? The Expected Value of Perfect Information (EVPI) answers this. EVPI is the difference between the expected value if you could know the uncertain variable perfectly *before* deciding and the expected value of the best decision made under uncertainty. If EVPI is large, the pilot is worth running -- the information will improve your decision enough to justify the delay. If EVPI is small, just pick the best current option and go. EVPI puts a hard ceiling on how much any study, survey, or pilot should cost.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch07-single-agent/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch07-single-agent/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

We build a decision analysis for Kenya's health ministry: three intervention strategies, uncertain government support, and a clear objective -- maximize neonatal lives saved.

### Set Up the Decision Problem

Three strategies, two scenarios, known probabilities, and estimated payoffs.

```{r}
#| label: decision-setup
#| echo: true

library(dplyr)

# Kenya decision problem: three strategies, two scenarios
strategies <- data.frame(
  strategy    = rep(c("Equipment Only", "Workforce", "Split"), each = 2),
  gov_support = rep(c("High", "Low"), 3),
  probability = rep(c(0.40, 0.60), 3),
  lives_saved = c(2800, 1200,    # Equipment: great with support, weak without
                  3600, 1000,    # Workforce: highest ceiling, lowest floor
                  3100, 1400))   # Split: balanced, less risky

# Compute expected value per strategy
ev_table <- strategies |>
  group_by(strategy) |>
  summarise(EV = sum(probability * lives_saved), .groups = "drop") |>
  arrange(desc(EV))

ev_table
```

### Calculate EVPI

How much is it worth to learn Kenya's government support level before committing?

```{r}
#| label: evpi-calc
#| echo: true

# Best decision in each scenario (if you knew the future)
best_per_scenario <- strategies |>
  group_by(gov_support) |>
  slice_max(lives_saved) |>
  ungroup()

# EV with perfect information
ev_perfect <- sum(best_per_scenario$probability * best_per_scenario$lives_saved)

# EV without information (best overall strategy)
ev_no_info <- max(ev_table$EV)

# EVPI = the value of knowing before you decide
evpi <- ev_perfect - ev_no_info
cat("EV with perfect info:", ev_perfect, "lives\n")
cat("EV best uninformed:  ", ev_no_info, "lives\n")
cat("EVPI:                ", evpi, "lives\n")
cat("\nIf a pilot costs less than", evpi, "lives worth of resources, fund it.")
```

### Sensitivity Analysis

How does the optimal strategy change as the probability of high government support shifts? The crossover points reveal where the optimal strategy flips.

```{r}
#| label: sensitivity-plot
#| echo: true
#| fig-width: 9
#| fig-height: 5

library(ggplot2)

# Vary P(High support) from 0 to 1
p_seq <- seq(0, 1, by = 0.01)

sens <- data.frame(
  p = rep(p_seq, 3),
  strategy = rep(c("Equipment Only", "Workforce", "Split"), each = length(p_seq)),
  ev = c(1200 + 1600 * p_seq,    # Equipment
         1000 + 2600 * p_seq,    # Workforce
         1400 + 1700 * p_seq))   # Split

ggplot(sens, aes(x = p, y = ev, color = strategy)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0.40, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 0.42, y = 2800, label = "Current estimate\np = 0.40",
           hjust = 0, size = 3.5) +
  scale_color_manual(values = c("#005CB9", "#83BD00", "#FA7650")) +
  labs(x = "P(High Government Support)",
       y = "Expected Lives Saved",
       title = "Sensitivity Analysis: Which Strategy Wins?",
       subtitle = "Crossover points show where the optimal strategy changes") +
  theme_minimal(base_size = 14)
```

### Graph Surgery: Watching vs. Doing

We build two versions of a simple network -- one with confounding, one with the arrow cut -- and show how the causal estimate differs from the observational one.

```{r}
#| label: graph-surgery
#| echo: true

library(bnlearn)

# Original model: Resources confounds both Equipment and NMR
original <- model2network(
  "[Resources][Equipment|Resources][NMR|Equipment:Resources]")

# After graph surgery: do(Equipment) -- cut Resources -> Equipment
intervened <- model2network(
  "[Resources][Equipment][NMR|Equipment:Resources]")

# Original: P(NMR | Equipment) includes confounding via Resources
# Intervened: P(NMR | do(Equipment)) is the true causal effect
# The difference is the confounding bias -- graph surgery removes it.
```

### Tornado Diagram

Which uncertain parameter swings the result the most? Parameters with wider bars deserve the most attention.

```{r}
#| label: tornado
#| echo: true
#| fig-width: 9
#| fig-height: 4

params <- data.frame(
  parameter = c("P(Gov Support)", "Lives if High", "Lives if Low"),
  low_ev  = c(1840, 1960, 1960),
  high_ev = c(2320, 2200, 2200))
params$swing <- params$high_ev - params$low_ev
params <- params[order(params$swing), ]
params$parameter <- factor(params$parameter, levels = params$parameter)

ggplot(params, aes(y = parameter)) +
  geom_segment(aes(x = low_ev, xend = high_ev, yend = parameter),
               linewidth = 8, color = "#005CB9", alpha = 0.7) +
  geom_vline(xintercept = 2080, linetype = "dashed") +
  labs(x = "Expected Lives Saved", y = "",
       title = "Tornado Diagram: What Matters Most for the Split Strategy?") +
  theme_minimal(base_size = 14)
```

## Key Takeaways

::: {.key-takeaway}
**Influence diagrams add decisions and objectives to causal models.** This is the bridge from understanding the world (inference) to changing it (intervention).

**Expected value gives you a principled way to choose under uncertainty.** You will not always get the best outcome, but you will make the smartest bet on average.

**Watching is not doing.** Observational correlations include confounding. The do-operator (graph surgery) isolates the true causal effect by cutting incoming arrows to the intervention variable.

**EVPI tells you the value of better information.** If the pilot costs more than the EVPI, skip it and decide now. If it costs less, the information is worth the wait.
:::

## Looking Ahead

Next in **Chapter 8 -- Multi-Country Resource Allocation**, we scale up from a single country decision to a portfolio problem:

- Allocating a fixed budget across Ethiopia, Rwanda, Kenya, and Tanzania simultaneously
- Sequential decisions -- pilot first, or scale immediately?
- Country-specific risk profiles and the tension between maximizing total impact and ensuring equity
