---
title: "Chapter 7 — From Understanding to Deciding"
subtitle: "David, one envelope, and one new midwife"
---

## The mystery

A county budget director in Nakuru has enough money for one new midwife. Two facilities are competing for her — one busy and urban, one rural and unstaffed. Both managers have written passionate letters. He needs an answer he can defend at the next public budget hearing — not a gut call.

::: {.callout-note}
**The one thing to remember from this chapter:** A decision is just a question with consequences — and the value of more information is exactly how much it would change your decision.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch07-decision-analysis/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch07-decision-analysis/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — compute David's expected value, then his EVPI

In the lecture, David weighed Bahati against Nakuru using natural-frequency expected values. In this workshop you'll do the same calculation in R, then compute the Expected Value of Perfect Information (EVPI) — the answer to "would more research actually change my decision?".

### Setting up

```{r}
#| label: ch07-setup
library(dplyr)
library(tibble)
```

### David's two options

```{r}
#| label: ch07-options
# Each row is one possible future for one option, with the lives saved and
# the probability of that future.
options <- tribble(
  ~option,        ~scenario,    ~lives_saved, ~prob,
  "Bahati",       "good year",  8,            0.60,
  "Bahati",       "bad year",   2,            0.40,
  "Nakuru town",  "good year",  5,            0.80,
  "Nakuru town",  "bad year",   1,            0.20
)

options
```

### Compute the expected value for each option

```{r}
#| label: ch07-ev
options |>
  group_by(option) |>
  summarise(
    expected_lives_saved = sum(lives_saved * prob)
  )
```

You should see Bahati at 5.6 and Nakuru town at 4.2. Bahati wins by 1.4 lives per year.

### Compute the Expected Value of Perfect Information

```{r}
#| label: ch07-evpi
# Step 1: under uncertainty (no perfect info), the best David can do is pick
# the option with the highest expected value.
ev_no_info <- options |>
  group_by(option) |>
  summarise(ev = sum(lives_saved * prob)) |>
  pull(ev) |>
  max()

# Step 2: with perfect information, David would know which scenario is going
# to happen *before* he chooses. So in each scenario, he can pick the BEST
# option for that scenario.
ev_with_info <- options |>
  group_by(scenario) |>
  summarise(
    best_lives = max(lives_saved),
    prob       = first(prob)
  ) |>
  summarise(ev = sum(best_lives * prob)) |>
  pull(ev)

# EVPI is the difference.
evpi <- ev_with_info - ev_no_info
evpi
```

What you should see: an EVPI of around 1 life per year. That tells David that a one-month study of Bahati's complication rate would, in the best case, save about one extra newborn life — which is meaningful but not enormous. He'd weigh that against the cost of waiting a month before placing the midwife.

This is the "would more research change my decision?" question, made into a number.

## Try it yourself

A donor offers your county a free newborn resuscitation device. Two hospitals want it. Hospital A is large and busy with reliable electricity (3 lives saved, high confidence). Hospital B is smaller and rural, with power cuts about half the time (6 lives saved when power holds, 1 when it doesn't). Compute the expected value for each. Where does the device go?
