---
title: "Chapter 7 — From Understanding to Deciding"
subtitle: "David, one envelope, and one new midwife"
---

## The mystery

A county budget director in Nakuru has enough money for one new midwife. Two facilities are competing for her — one busy and urban, one rural and unstaffed. Both managers have written passionate letters. He needs an answer he can defend at the next public budget hearing — not a gut call.

::: {.callout-note}
**The one thing to remember from this chapter:** A decision is just a question with consequences — and the value of more information is exactly how much it would change your decision.
:::

## Main lecture

::: {.slide-embed}
<iframe src="../ch07-single-agent/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch07-single-agent/index.html" target="_blank">Open lecture slides full-screen &rarr;</a>
</div>
:::

## Tutorial / breakout session

The tutorial deck takes the same ideas and retells them as a story — built for a live breakout session, not a lecture.

::: {.slide-embed}
<iframe src="../ch07-decision-analysis-tutorial/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch07-decision-analysis-tutorial/index.html" target="_blank">Open tutorial slides full-screen &rarr;</a>
</div>
:::

## R workshop — compute David's expected value, then his EVPI

In the lecture, David weighed Bahati against Nakuru using natural-frequency expected values. In this workshop you'll do the same calculation in R, then compute the Expected Value of Perfect Information (EVPI) — the answer to "would more research actually change my decision?".

### Setting up

```{r}
#| label: ch07-setup
library(dplyr)
library(tibble)
```

### David's two options

```{r}
#| label: ch07-options
# Each row is one possible future for one option, with the lives saved and
# the probability of that future.
options <- tribble(
  ~option,        ~scenario,    ~lives_saved, ~prob,
  "Bahati",       "good year",  8,            0.60,
  "Bahati",       "bad year",   2,            0.40,
  "Nakuru town",  "good year",  5,            0.80,
  "Nakuru town",  "bad year",   1,            0.20
)

options
```

### Compute the expected value for each option

```{r}
#| label: ch07-ev
options |>
  group_by(option) |>
  summarise(
    expected_lives_saved = sum(lives_saved * prob)
  )
```

You should see Bahati at 5.6 and Nakuru town at 4.2. Bahati wins by 1.4 lives per year.

### Compute the Expected Value of Perfect Information

EVPI requires us to enumerate the *joint* scenarios across both options, not just one. We assume Bahati's good/bad year is independent of Nakuru's good/bad year — a rough but defensible assumption when the two facilities are far enough apart that local shocks (rainfall, road closures, staff turnover) don't move together.

```{r}
#| label: ch07-evpi
# Bahati and Nakuru as two separate uncertain sources.
bahati_states <- tibble(state = c("good", "bad"), lives = c(8, 2), prob = c(0.6, 0.4))
nakuru_states <- tibble(state = c("good", "bad"), lives = c(5, 1), prob = c(0.8, 0.2))

# All four joint states (Bahati state x Nakuru state).
joint <- tidyr::crossing(
  b_state = bahati_states$state,
  n_state = nakuru_states$state
) |>
  mutate(
    b_lives = bahati_states$lives[match(b_state, bahati_states$state)],
    n_lives = nakuru_states$lives[match(n_state, nakuru_states$state)],
    b_prob  = bahati_states$prob[match(b_state, bahati_states$state)],
    n_prob  = nakuru_states$prob[match(n_state, nakuru_states$state)],
    joint_prob = b_prob * n_prob,
    # In each joint state, with perfect info you'd pick the option that
    # actually saves more lives in that state.
    best_lives = pmax(b_lives, n_lives)
  )

joint

# EV under uncertainty: David picks the single option with the highest
# expected value (Bahati at 5.6) and lives with whatever year actually happens.
ev_no_info <- max(
  sum(bahati_states$lives * bahati_states$prob),
  sum(nakuru_states$lives * nakuru_states$prob)
)
ev_no_info

# EV with perfect information: in each joint state, David picks the best
# option for that state.
ev_with_info <- sum(joint$best_lives * joint$joint_prob)
ev_with_info

# EVPI = how much you'd gain on average from knowing the future before deciding.
evpi <- ev_with_info - ev_no_info
evpi
```

What you should see: `ev_no_info` is 5.6 (Bahati's expected value), `ev_with_info` is around 6.56, and **`evpi` is around 0.96 — about 1 extra life per year**. That tells David a perfectly informative one-month study could, in expectation, save him about one extra newborn life — which is meaningful but not enormous. He'd weigh that against the cost of waiting a month before placing the midwife.

This is the "would more research change my decision?" question, made into a number.

## Try it yourself

A donor offers your county a free newborn resuscitation device. Two hospitals want it. Hospital A is large and busy with reliable electricity (3 lives saved, high confidence). Hospital B is smaller and rural, with power cuts about half the time (6 lives saved when power holds, 1 when it doesn't). Compute the expected value for each. Where does the device go?
