---
title: "Chapter 8: Multi-Country Resource Allocation"
subtitle: "How do you split a limited budget across four countries to save the most lives?"
---

## Learning Objectives

::: {.learning-objectives}
- Explain why **allocating resources across countries** is harder than picking the single best one
- Compare three allocation strategies: **Concentrate**, **Spread Equally**, and **Tiered**
- Use **Monte Carlo simulation** to see how uncertainty changes the best strategy
- Understand **sequential decisions** -- why you might pilot first and scale later
- Run a Monte Carlo simulation in R and interpret the results
:::

## Key Concepts

::: {.key-concept}
#### The Allocation Puzzle

Imagine you run a maternal and newborn health program with a fixed budget and four countries to support -- say Ethiopia, Tanzania, Rwanda, and Nigeria. Each country has different population sizes, different health system strengths, and different levels of government commitment. You cannot simply rank them and fund the top one, because spreading your investment might reduce the risk that a single country's problems sink your entire program.
:::

::: {.key-concept .green}
#### Three Strategies: Concentrate, Spread, or Tier

**Concentrate** means putting most of your budget into the country where you expect the biggest impact per dollar. High reward, but high risk -- if that country hits political trouble or a supply chain crisis, you lose big. **Spread** means dividing the budget equally across all four countries. Safer, but you might waste money in countries where it does less good. **Tiered** means weighting the budget by each country's burden and capacity, with minimum and maximum caps per country. This balances efficiency and safety.
:::

::: {.key-concept .orange}
#### Uncertainty and Monte Carlo Simulation

You never know exactly how effective your spending will be. Government co-financing might be strong or weak. Drug supply chains might hold up or break down. Monte Carlo simulation handles this by running your model thousands of times, each time drawing random values for the uncertain inputs. Instead of a single "expected lives saved" number, you get a whole distribution -- and you can ask questions like "What is the chance we save fewer than 5,000 lives?"
:::

::: {.key-concept .red}
#### Sequential Decisions: Pilot, Then Scale

You do not have to commit your entire budget on day one. A smarter approach is to invest a small amount first (a pilot), observe results, and then scale up funding in the countries that performed well. This is called **sequential decision making**. The catch? Every month you spend gathering information is a month when people who could have been helped are not getting help. So there is a real trade-off between learning and acting.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch08-resource-allocation/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch08-resource-allocation/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

We will build a Monte Carlo simulation to compare allocation strategies across four countries. The idea is simple: define what you know (and what you are uncertain about), then let the computer run thousands of "what if" scenarios.

### Setting Up Country Parameters

Each country has three numbers: how many lives per million dollars you expect to save, how uncertain that estimate is (the standard deviation), and how likely the government is to provide strong co-financing.

```{r}
#| label: setup-countries
#| echo: true

set.seed(42)
n_sims <- 10000

# Four countries with public-data-inspired parameters
countries <- data.frame(
  country    = c("Ethiopia", "Tanzania", "Rwanda", "Nigeria"),
  lives_per_M = c(610, 462, 356, 520),    # expected lives saved per $1M
  sd_per_M    = c(185, 130, 95, 160),      # uncertainty in effectiveness
  p_high_gov  = c(0.40, 0.45, 0.50, 0.30) # P(strong gov co-financing)
)

countries
```

### The Simulation Function

This function takes a budget allocation (four numbers that add to 100) and runs `n_sims` random scenarios. In each scenario, it draws a government co-financing level and a random effectiveness for each country, then totals the lives saved.

```{r}
#| label: simulate-function
#| echo: true

simulate_outcomes <- function(alloc, countries, n_sims) {
  total_lives <- numeric(n_sims)
  for (i in 1:n_sims) {
    lives <- 0
    for (j in 1:nrow(countries)) {
      # Draw a government co-financing scenario
      gov <- sample(c("High", "Med", "Low"), 1,
                    prob = c(countries$p_high_gov[j],
                             1 - countries$p_high_gov[j] - 0.10,
                             0.10))
      multiplier <- switch(gov, High = 1.0, Med = 0.6, Low = 0.25)

      # Lives saved = allocation share * random effectiveness * multiplier
      lives_j <- alloc[j] *
        rnorm(1, countries$lives_per_M[j], countries$sd_per_M[j]) *
        multiplier
      lives <- lives + max(0, lives_j)
    }
    total_lives[i] <- lives
  }
  total_lives
}
```

### Comparing Three Strategies

Now we define our three strategies and simulate 10,000 outcomes for each. The violin plot shows the full spread of possible results -- not just the average, but the best and worst cases too.

```{r}
#| label: compare-strategies
#| echo: true
#| fig-width: 10
#| fig-height: 5

library(ggplot2)

# Three allocation strategies (% of budget to each country)
#                     Ethiopia  Tanzania  Rwanda  Nigeria
strat_concentrate <- c(55,       15,       10,     20)
strat_spread      <- c(25,       25,       25,     25)
strat_tiered      <- c(35,       25,       15,     25)

# Run simulations
sim_conc   <- simulate_outcomes(strat_concentrate, countries, n_sims)
sim_spread <- simulate_outcomes(strat_spread,      countries, n_sims)
sim_tiered <- simulate_outcomes(strat_tiered,      countries, n_sims)

# Combine into a data frame for plotting
sim_df <- data.frame(
  lives    = c(sim_conc, sim_spread, sim_tiered),
  strategy = rep(c("Concentrate", "Spread", "Tiered"), each = n_sims)
)

ggplot(sim_df, aes(x = strategy, y = lives / 1000, fill = strategy)) +
  geom_violin(alpha = 0.6, draw_quantiles = c(0.05, 0.5, 0.95)) +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.shape = NA) +
  scale_fill_manual(values = c("#005CB9", "#83BD00", "#FA7650")) +
  labs(x = "Allocation Strategy",
       y = "Total Lives Saved (thousands)",
       title = "Monte Carlo Simulation: 4-Country Allocation",
       subtitle = "10,000 scenarios | Lines show 5th, 50th, and 95th percentiles") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

### Sensitivity: What If Government Support Changes?

The biggest unknown is usually whether a government will step up its own spending. This plot shows how each strategy performs as Ethiopia's probability of strong co-financing varies from 10% to 80%.

```{r}
#| label: sensitivity
#| echo: true
#| fig-width: 10
#| fig-height: 5

library(tidyr)

p_seq <- seq(0.1, 0.8, by = 0.05)
sens_results <- data.frame()

for (p_high in p_seq) {
  countries_mod <- countries
  countries_mod$p_high_gov[1] <- p_high  # vary Ethiopia

  ev_conc <- mean(simulate_outcomes(strat_concentrate, countries_mod, 3000))
  ev_spr  <- mean(simulate_outcomes(strat_spread,      countries_mod, 3000))
  ev_tier <- mean(simulate_outcomes(strat_tiered,       countries_mod, 3000))

  sens_results <- rbind(sens_results, data.frame(
    p_high = p_high,
    Concentrate = ev_conc, Spread = ev_spr, Tiered = ev_tier
  ))
}

sens_long <- pivot_longer(sens_results, cols = -p_high,
                          names_to = "Strategy", values_to = "EV")

ggplot(sens_long, aes(x = p_high, y = EV / 1000, color = Strategy)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0.40, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("#005CB9", "#83BD00", "#FA7650")) +
  labs(x = "P(Strong Gov Co-financing) for Ethiopia",
       y = "Expected Lives Saved (thousands)",
       title = "How Ethiopia's Government Commitment Changes the Best Strategy",
       subtitle = "Dashed line = our current best guess (p = 0.40)") +
  theme_minimal(base_size = 14)
```

Notice how the "Concentrate" line is steepest -- it benefits the most when Ethiopia's government steps up, but it also falls the hardest when support is weak. The "Tiered" strategy is more robust: it does reasonably well no matter what happens.

## Key Takeaways

::: {.key-takeaway}
**Allocating resources across countries** is not just about picking the most efficient one. Uncertainty means the "best" country might disappoint, so diversification reduces catastrophic risk -- just like in a financial portfolio.

**Monte Carlo simulation** lets you explore thousands of possible futures instead of relying on a single "expected" number. It reveals the full range of outcomes, including worst-case scenarios that matter for real decisions.

**Sequential decisions** (pilot first, then scale) are valuable because they let you learn before committing. But delay has a cost measured in lives, so you need to balance the value of information against the urgency of action.
:::

## Looking Ahead

Next in **Chapter 9 -- Game Theory and Strategic Interactions**, we stop treating governments and other donors as random variables and start treating them as strategic players with their own goals:

- What happens when multiple players each try to maximize their own payoff?
- Why do rational players sometimes end up in outcomes that are bad for everyone?
- How can you design contracts and incentives that make cooperation the smart choice?

Chapter 8 asked "How should we split the budget?" assuming other actors are unpredictable. Chapter 9 recognizes that those actors are making rational choices -- and shows how to design strategies that work *with* their incentives rather than against them.
