---
title: "Chapter 8 — Spreading the Money"
subtitle: "Grace, four countries, and three risk shapes"
---

## The mystery

A programme officer has a fixed MNH budget and four countries. Her board meets in three weeks and wants one slide. Concentrate? Spread? Tier? Each strategy has a different *shape of risk* — and the average expected value alone cannot pick a winner.

::: {.callout-note}
**The one thing to remember from this chapter:** Spreading resources is not a math question — it's a question about which uncertainties you're willing to live with. The expected value picks a centre. The shape of the range picks a strategy.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch08-resource-allocation/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch08-resource-allocation/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — Monte Carlo Grace's three strategies

In the lecture, Grace compared concentrate, spread, and tier on expected value alone — and saw that the expected values didn't pick a winner. In this workshop you'll run a Monte Carlo simulation for each strategy in R and *see the shapes of the three histograms* directly.

### Setting up

```{r}
#| label: ch08-setup
library(dplyr)
library(ggplot2)
library(tidyr)

set.seed(42)  # so the random draws are reproducible
```

### The four countries

```{r}
#| label: ch08-countries
# Per-quarter-budget estimates from the lecture table.
# Each country has a "low" and "high" estimate; we treat the truth as
# uniformly distributed between them — a rough but defensible assumption.
countries <- tibble::tribble(
  ~country,    ~low,  ~high,
  "Ethiopia",   600,  1800,
  "Rwanda",     700,  1100,
  "Kenya",      400,  1600,
  "Tanzania",   500,  1700
)

countries
```

### Simulate one possible future

```{r}
#| label: ch08-one-future
# For one possible future, draw one outcome per country.
# (Quarter-budget effect — we'll scale by allocation share below.)
draw_one_future <- function() {
  countries |>
    mutate(outcome = runif(n(), low, high))
}

draw_one_future()
```

### Run 10,000 futures for each strategy

```{r}
#| label: ch08-strategies
# Concentrate: all the money on Ethiopia (highest expected value).
# Scale-up factor of 4 with diminishing returns: a 4x budget gets you ~3x outcome.
concentrate_strategy <- function() {
  draw <- runif(1, 600, 1800)  # Ethiopia draw
  draw * 3                       # diminishing-returns scale-up
}

# Spread: each country gets one quarter — sum the four outcomes.
spread_strategy <- function() {
  futures <- draw_one_future()
  sum(futures$outcome)
}

# Tier: 40% Ethiopia, 40% Tanzania (top two), 10% Rwanda, 10% Kenya.
tier_strategy <- function() {
  futures <- draw_one_future()
  weights <- c(Ethiopia = 1.6, Rwanda = 0.4, Kenya = 0.4, Tanzania = 1.6)
  sum(futures$outcome * weights[futures$country])
}

# Run each strategy 10,000 times.
n_sims <- 10000
sims <- tibble(
  Concentrate = replicate(n_sims, concentrate_strategy()),
  Spread      = replicate(n_sims, spread_strategy()),
  Tier        = replicate(n_sims, tier_strategy())
)

summary(sims)
```

### Compare the three histograms

```{r}
#| label: ch08-histograms
sims_long <- sims |>
  pivot_longer(everything(), names_to = "strategy", values_to = "lives_saved")

ggplot(sims_long, aes(x = lives_saved, fill = strategy)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 50) +
  facet_wrap(~strategy, ncol = 1) +
  labs(
    title    = "Three strategies, three risk shapes",
    subtitle = "Each histogram is 10,000 simulated futures",
    x        = "Lives saved",
    y        = "Number of futures"
  ) +
  theme_minimal()
```

What you should see:

- **Concentrate** — a wide histogram with a long left tail. Best case is huge; worst case is the worst of any strategy.
- **Spread** — a fat blob in the middle. Almost no futures are great; almost no futures are terrible.
- **Tier** — a narrower bell shifted right of spread, with a higher floor than concentrate.

This is the conversation Grace needs to have with her board — not "whose number is bigger" but "which shape of risk are we willing to live with".

### Answer the board's real question

```{r}
#| label: ch08-board-question
# What fraction of futures end up under 3000 lives saved?
sims_long |>
  group_by(strategy) |>
  summarise(
    pct_under_3000 = mean(lives_saved < 3000)
  )
```

That's the number that decides the board meeting. *"There is a 12% chance of ending up under 3000 lives saved"* is a sentence a board member can argue with. *"The expected value is 4500"* is not.

## Try it yourself

You have a fixed MNH budget. Three options: (A) a national rollout in a country with strong recent results but moderate uncertainty; (B) a regional pilot in a country with weak recent results but very high upside if it works; (C) a 50/50 split. Sketch the three risk shapes without computing anything. Which one would you bring to the board, and why?
