---
title: "Chapter 9: Game Theory & Strategic Interactions"
subtitle: "When everyone acts in their own interest, how do you get cooperation across Kenya's 47 counties?"
---

## Learning Objectives

::: {.learning-objectives}
- Explain what a **game** is in the formal sense: players, strategies, and payoffs
- Read and interpret a **payoff matrix** to find each player's best response
- Find the **Nash Equilibrium** -- the outcome where nobody wants to change their move
- Recognize the **free-rider problem** and explain why it leads to bad outcomes
- Describe how **commitment devices** (like conditional grants) can fix the problem
:::

## Key Concepts

::: {.key-concept}
#### Strategic Interactions

In Chapter 8, we treated government co-financing as a random variable -- something uncertain but not strategic. In reality, county governments are rational actors making deliberate choices. When a large donor announces generous funding for Kenya's health sector, a county government might think: "Great, we can redirect our own health budget to roads." This is not random -- it is a strategic response. Kenya's devolved system splits authority across 47 counties, each with its own health budget, making these strategic calculations happen at every level. Game theory gives us the tools to anticipate and shape these responses.
:::

::: {.key-concept .green}
#### Payoff Matrices

A payoff matrix is a table that shows what happens for every possible combination of choices. Imagine a donor and a Kenya county government each have two options: invest heavily or invest lightly in newborn health. The matrix has four cells, one for each combination. Each cell contains two numbers: the donor's payoff and the county's payoff. Reading the matrix tells you what each player should do given what the other player does.
:::

::: {.key-concept .orange}
#### Nash Equilibrium

A Nash Equilibrium is a combination of strategies where neither player wants to change their choice, given what the other player is doing. Think of it as the "stable" outcome -- once both players land there, nobody has a reason to move. The problem is that the Nash Equilibrium is not always the best outcome for everyone. In many health funding games, the equilibrium involves the donor paying generously while the county government free-rides.
:::

::: {.key-concept .red}
#### The Free-Rider Problem

Free-riding happens when one player benefits from the other's effort without contributing their fair share. If the donor will fund a Kenyan county's neonatal health program regardless of what the county does, the county's best move is to spend its money elsewhere. Some of Kenya's 47 counties redirect conditional health grants toward other priorities while still claiming credit for donor-funded health improvements. This is individually rational but collectively harmful -- total investment in health ends up lower than it should be.
:::

::: {.key-concept .teal}
#### Commitment Devices

A commitment device is a contract structure that changes the payoff matrix so that cooperation becomes the smart choice for both sides. Kenya's conditional grant mechanism is a real-world example: the national government commits funding, but each tranche is released only when a county hits specific health targets verified through DHIS2 (like reducing neonatal mortality or increasing facility deliveries). This makes free-riding costly and cooperation genuinely attractive. The key is credibility -- the national government or donor must actually withhold funds when milestones are missed, or the threat is empty.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch09-multi-agent/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch09-multi-agent/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

We will build a payoff matrix in R, find the Nash Equilibrium, and then simulate a repeated game to see how tit-for-tat can induce cooperation between a donor and a county government.

### Building a Payoff Matrix and Finding Nash Equilibrium

Each player can "Invest High" or "Invest Low" in neonatal health. The donor cares about lives saved; the county government balances health against other priorities like roads and water.

```{r}
#| label: payoff-and-nash
#| echo: true

# Donor payoffs (lives saved, in hundreds)
donor_payoffs <- matrix(
  c(24, 18, 5, 8), nrow = 2, ncol = 2,
  dimnames = list(Donor = c("Invest High", "Invest Low"),
                  County = c("Invest High", "Invest Low")))

# County government payoffs (composite score)
gov_payoffs <- matrix(
  c(7, 5, 10, 6), nrow = 2, ncol = 2,
  dimnames = list(Donor = c("Invest High", "Invest Low"),
                  County = c("Invest High", "Invest Low")))

# Find Nash Equilibria by checking best responses
find_nash <- function(p1, p2) {
  nash <- list()
  for (i in 1:nrow(p1))
    for (j in 1:ncol(p1))
      if (p1[i, j] == max(p1[, j]) & p2[i, j] == max(p2[i, ]))
        nash <- c(nash, list(c(row = i, col = j)))
  nash
}

ne <- find_nash(donor_payoffs, gov_payoffs)
for (eq in ne)
  cat(sprintf("Nash Equilibrium: (%s, %s) -> Donor: %d, County: %d\n",
              rownames(donor_payoffs)[eq[1]],
              colnames(donor_payoffs)[eq[2]],
              donor_payoffs[eq[1], eq[2]], gov_payoffs[eq[1], eq[2]]))
# Expected: (Invest High, Invest Low) -- the free-rider outcome
```

The equilibrium is (Invest High, Invest Low) -- the donor funds generously and the county government free-rides. Both could do better by cooperating, but neither has an individual reason to change.

### Simulating a Repeated Game with Tit-for-Tat

In a one-shot game, free-riding wins. But over 20 rounds, a tit-for-tat donor -- who matches the county's previous move -- can punish free-riding and reward cooperation. Think of each round as a year of conditional grant disbursements.

```{r}
#| label: repeated-game
#| echo: true
#| fig-width: 10
#| fig-height: 5

library(ggplot2)
set.seed(42)

simulate_repeated <- function(donor_strategy, n = 20) {
  d_act <- g_act <- character(n)
  d_pay <- g_pay <- numeric(n)
  payoffs <- list("High_High" = c(D=24,G=7), "High_Low" = c(D=5,G=10),
                  "Low_High"  = c(D=18,G=5), "Low_Low"  = c(D=8,G=6))
  for (t in 1:n) {
    d_act[t] <- if (donor_strategy == "always_high") "High"
                else if (t == 1) "High"
                else if (g_act[t-1] == "High") "High" else "Low"
    g_act[t] <- if (t <= 3) "Low"
                else if (donor_strategy == "tit_for_tat" & d_act[t] == "Low") "High"
                else "Low"
    p <- payoffs[[paste0(d_act[t], "_", g_act[t])]]
    d_pay[t] <- p["D"]; g_pay[t] <- p["G"]
  }
  data.frame(round=1:n, county_action=g_act, donor_payoff=d_pay)
}

always <- simulate_repeated("always_high"); always$strategy <- "Always Fund High"
tft    <- simulate_repeated("tit_for_tat"); tft$strategy    <- "Tit-for-Tat"
results <- rbind(always, tft)

ggplot(results, aes(x=round, y=cumsum(donor_payoff), color=strategy)) +
  geom_line(linewidth=1.2) +
  geom_point(aes(shape=county_action), size=2.5) +
  scale_color_manual(values=c("#E24A3F","#83BD00")) +
  scale_shape_manual(values=c(High=16, Low=4), name="County Action") +
  labs(title="Repeated Game: Does Conditional Funding Work?",
       subtitle="Tit-for-tat punishes free-riding and rewards cooperation",
       x="Funding Round (Year)", y="Cumulative Lives Saved (hundreds)",
       color="Donor Strategy") +
  theme_minimal(base_size=14) +
  theme(plot.title=element_text(color="#005CB9", face="bold"))
```

The unconditional donor gets exploited every round. The tit-for-tat donor takes a short-term hit when it punishes, but cooperation kicks in and cumulative lives saved climb faster.

### Payoff Matrix Heatmap

A heatmap shows which outcomes are good for the donor (green) and which are bad (red).

```{r}
#| label: payoff-heatmap
#| echo: true
#| fig-width: 9
#| fig-height: 5

game_df <- expand.grid(Donor=c("Invest High","Invest Low"),
                       County=c("Invest High","Invest Low"))
game_df$Donor_Payoff <- c(24,18,5,8)
game_df$County_Payoff <- c(7,5,10,6)
game_df$Label <- paste0("Donor: ",game_df$Donor_Payoff,
                        "\nCounty: ",game_df$County_Payoff)

ggplot(game_df, aes(x=County, y=Donor, fill=Donor_Payoff)) +
  geom_tile(color="white", linewidth=2) +
  geom_text(aes(label=Label), size=5, fontface="bold") +
  scale_fill_gradient2(low="#E24A3F", mid="#FED141", high="#83BD00",
                       midpoint=14, name="Donor\nPayoff") +
  labs(title="Kenya County Health Funding Game: Payoff Matrix",
       x="County Government Strategy", y="Donor Strategy") +
  theme_minimal(base_size=16) + theme(panel.grid=element_blank())
```

## Key Takeaways

::: {.key-takeaway}
**Game theory** helps you think about what happens when multiple players each pursue their own interests. In Kenya's devolved health system, this often means donors or the national government pay and county governments redirect money elsewhere.

**Nash Equilibrium** is the stable outcome, but it is not always the best outcome. The free-rider equilibrium wastes resources and costs lives.

**Commitment devices** like conditional grants change the payoff matrix so that cooperation becomes each player's best response. In a repeated game, tit-for-tat strategies can build cooperation over time -- but only if the threat of punishment is credible. Kenya's conditional grant mechanism, verified through DHIS2, is exactly this kind of device.
:::

## Looking Ahead

Next in **Chapter 10 -- Structure Learning from Data**, we ask whether computers can discover causal relationships on their own:

- Can an algorithm look at health data from Tanzania and figure out what causes what?
- Where do algorithms succeed, and where do they need human experts to fill in the gaps?
- How do you combine data-driven discovery with the expert knowledge we have been building all course?

Chapter 9 showed how to think about strategic actors. Chapter 10 returns to the data and asks the ultimate question: can we let the evidence tell us the causal structure, or do we always need a human in the loop?
