---
title: "Chapter 9 — When the Other Side Reacts"
subtitle: "Wangari, Patrick, and the same offer"
---

## The mystery

A donor offers two Kenyan counties the same matching grant for maternal health. Wangari accepts enthusiastically and doubles her maternal health budget. Patrick quietly redirects his maternal health funds to road repairs, knowing the donor will fund maternal health anyway. Same deal. Opposite responses. The donor's expected-value forecast was wrong in one of the two counties — and the *reason* it was wrong is the most important lesson in Part 3.

::: {.callout-note}
**The one thing to remember from this chapter:** Your decisions change how other people behave — and unless you write that down, your forecast of impact will be wrong.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch09-game-theory/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch09-game-theory/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — find the Nash equilibrium of the matching grant

In the lecture, we walked through the four-cell payoff matrix for the matching grant and found by hand that the only stable cell was the worst one. In this workshop you'll build the matrix in R, ask it for the Nash equilibrium directly, then change one number and watch the equilibrium shift.

### Setting up

```{r}
#| label: ch09-setup
library(dplyr)
```

### Build the payoff matrix

```{r}
#| label: ch09-payoffs
# We'll represent the game as a small data frame.
# Each row is one cell of the matrix. Payoffs are in arbitrary units —
# higher = better for that player.
game <- tibble::tribble(
  ~county_action,    ~donor_action,    ~county_payoff, ~donor_payoff,
  "fund",            "match",          5,              5,
  "fund",            "walk_away",      2,              0,
  "redirect",        "match",          7,              -2,
  "redirect",        "walk_away",      4,              0
)

game
```

The payoffs reflect the lecture's reasoning:

- `fund + match` is good for both — but the county is leaving roads unrepaired.
- `redirect + match` is *best* for the county (gets roads + match) but *worst* for the donor (money wasted).
- `redirect + walk_away` is the cautious outcome — county gets roads, donor saves face.

### Find the Nash equilibrium by hand

A cell is a Nash equilibrium if neither player would improve by switching unilaterally.

```{r}
#| label: ch09-nash
find_nash <- function(g) {
  g |>
    rowwise() |>
    mutate(
      # Best response by the COUNTY, holding the donor's choice fixed
      county_best = max(g$county_payoff[g$donor_action == donor_action]),
      county_at_best = (county_payoff == county_best),

      # Best response by the DONOR, holding the county's choice fixed
      donor_best = max(g$donor_payoff[g$county_action == county_action]),
      donor_at_best = (donor_payoff == donor_best),

      is_nash = county_at_best & donor_at_best
    ) |>
    ungroup()
}

find_nash(game) |>
  select(county_action, donor_action, county_payoff, donor_payoff, is_nash)
```

You should see exactly one row with `is_nash = TRUE` — the bottom-right cell, where the county redirects and the donor walks away. **The rational equilibrium is the worst outcome for everyone.** That's the lecture's point in numbers.

### Now fix the game with conditionality

Suppose the donor adds a clause: *"the match only triggers if total maternal health spending rises by at least X."* This makes `redirect + match` impossible — its payoff drops sharply for the county.

```{r}
#| label: ch09-fixed
game_fixed <- game |>
  mutate(county_payoff = case_when(
    county_action == "redirect" & donor_action == "match" ~ -1,  # large penalty
    TRUE ~ county_payoff
  ))

find_nash(game_fixed) |>
  select(county_action, donor_action, county_payoff, donor_payoff, is_nash)
```

The Nash equilibrium has moved. Now `fund + match` is the stable outcome — exactly what the donor wanted in the first place. **Same players, same actions, different payoffs, different equilibrium.**

That single shift — adding a single constraint — is what game-theory-aware grant design buys you.

## Try it yourself

A donor wants to launch a free emergency obstetric kit programme in three districts. One district uses the kits as intended; another sells them to neighbouring private clinics; a third refuses entirely. Draw a payoff matrix for the second district. What did the donor's incentive structure miss? Which fix — conditionality, commitment device, or joint review — would you use?
