---
title: "Chapter 10: Structure Learning from Data"
subtitle: "Can computers discover what causes what? Lessons from Tanzania's health data."
---

## Learning Objectives

::: {.learning-objectives}
- Explain what **structure learning** means and why it matters for causal analysis
- Describe what algorithms **can and cannot** discover from observational data alone
- Use the **hill-climbing algorithm** (`bnlearn::hc()`) in R to learn a DAG from data
- Understand why **expert knowledge + algorithms** beats either one alone
- Recognize the **feedback loop** between data, models, and decisions
:::

## Key Concepts

::: {.key-concept}
#### Can Computers Discover Causes?

Throughout this course, we have built causal models by hand -- drawing DAGs based on expert knowledge about how health systems work. But what if you have a large dataset from Tanzania's health information system (DHIS2) and you want the data itself to suggest which variables cause which? Structure learning algorithms try to do exactly that. They look at patterns of correlation and conditional independence in the data and propose a DAG that could have generated those patterns.
:::

::: {.key-concept .green}
#### What Algorithms Can and Cannot Do

Here is the honest truth: algorithms are powerful but limited. They **can** discover which variables are connected and, in some cases, which direction the arrows point. They **cannot** distinguish between all possible causal structures. Multiple different DAGs can produce the exact same statistical patterns -- this is called **Markov equivalence**. When the algorithm says "I found a link between staffing and mortality," it might not be able to tell you whether staffing causes lower mortality or whether something else drives both. That ambiguity is where human experts come in.
:::

::: {.key-concept .orange}
#### Expert + Algorithm: Better Together

The best results come from combining what the algorithm finds in the data with what domain experts know about the real world. The algorithm might discover a strong link between drug availability and neonatal outcomes in Tanzania. An expert can then confirm the direction (drugs help patients, not the other way around) and flag links the algorithm might have missed (like seasonal rainfall affecting both supply chains and disease burden). Think of the algorithm as a research assistant that does the grunt work, while the expert provides the judgment.
:::

::: {.key-concept .red}
#### The Feedback Loop

Structure learning is not a one-time event. It is part of a cycle: (1) collect data from facilities, (2) run the algorithm to discover candidate structures, (3) have experts review and refine the DAG, (4) use the refined model to make intervention decisions, (5) collect new data after the intervention, and (6) update the model. Each trip around this loop makes the model better. This is how a health program in Tanzania (or anywhere) gets smarter over time.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch10-data-driven/index.html" width="100%" height="500px" style="border: 1px solid #ddd; border-radius: 8px;"></iframe>
<div class="slide-embed-footer">
<a href="../ch10-data-driven/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

We will simulate a dataset that looks like Tanzania's DHIS2 facility data, run the hill-climbing structure learning algorithm, and compare what the algorithm discovers to the true causal structure we built into the simulation.

### Simulating Facility-Level Data

We create a dataset with a known causal structure so we can check how well the algorithm recovers it. The variables represent real health system factors: training, drug availability, staffing, equipment, delivery mode, referral patterns, patient volume, and neonatal mortality rate.

```{r}
#| label: simulate-data
#| echo: true

library(bnlearn)
library(tibble)

set.seed(42)
n <- 2000  # facilities in Tanzania

# The TRUE causal structure (we know this; the algorithm does not)
true_dag <- model2network(paste0(
  "[Training][DrugAvail][Equipment]",
  "[Staffing|Training]",
  "[CSection|Staffing:DrugAvail:Equipment]",
  "[Referral|Staffing:Equipment]",
  "[Volume|Staffing:Equipment]",
  "[NMR|CSection:Referral:Staffing:DrugAvail]"
))

# Generate data from the true structure
training    <- rbinom(n, 1, 0.55)
drug_avail  <- rbinom(n, 1, 0.65)
equipment   <- rnorm(n, 0.5, 0.15) |> pmin(1) |> pmax(0)
staffing    <- rbinom(n, 1, 0.35 + 0.30 * training)
csection    <- 5 + 8 * staffing + 3 * drug_avail +
               4 * equipment + rnorm(n, 0, 3)
referral    <- 15 - 5 * staffing - 3 * equipment + rnorm(n, 0, 4)
volume      <- 50 + 30 * staffing + 20 * equipment + rnorm(n, 0, 15)
nmr         <- 35 - 8 * staffing - 4 * drug_avail -
               3 * (csection > 10) - 2 * (referral < 10) +
               rnorm(n, 0, 5)

tz_data <- tibble(
  Training  = factor(training, labels = c("No", "Yes")),
  DrugAvail = factor(drug_avail, labels = c("Stockout", "Available")),
  Equipment = equipment,
  Staffing  = factor(staffing, labels = c("Low", "Adequate")),
  CSection  = csection,
  Referral  = referral,
  Volume    = volume,
  NMR       = nmr
)

head(tz_data)
```

### Running the Hill-Climbing Algorithm

The `hc()` function searches over possible DAG structures and picks the one that best balances fitting the data and keeping the model simple (using the BIC score). It is like trying every possible arrangement of arrows and keeping the one that scores best.

```{r}
#| label: hill-climbing
#| echo: true

# Score-based structure learning: Hill Climbing with BIC
learned_dag <- hc(tz_data, score = "bic-cg")
# "bic-cg" handles mixed data (both factors and continuous variables)

cat("Edges the algorithm found:\n")
print(arcs(learned_dag))

cat("\nEdges in the true DAG:\n")
print(arcs(true_dag))

# How close did it get? (Structural Hamming Distance -- lower is better)
shd_score <- shd(learned_dag, true_dag)
cat("\nStructural Hamming Distance:", shd_score, "\n")
cat("(0 = perfect recovery; higher = more mistakes)\n")
```

### Comparing True vs. Learned Structures

Let's put the true DAG and the learned DAG side by side. Where they agree, the algorithm got it right. Where they differ, we need expert judgment to decide which version is correct.

```{r}
#| label: compare-dags
#| echo: true
#| fig-width: 10
#| fig-height: 5

# Count correct, extra, and missing edges
true_arcs    <- arcs(true_dag)
learned_arcs <- arcs(learned_dag)
correct      <- merge(true_arcs, learned_arcs)

cat("=== Algorithm Report Card ===\n")
cat("True edges:       ", nrow(true_arcs), "\n")
cat("Learned edges:    ", nrow(learned_arcs), "\n")
cat("Correctly found:  ", nrow(correct), "\n")
cat("Extra (false):    ", nrow(learned_arcs) - nrow(correct), "\n")
cat("Missed:           ", nrow(true_arcs) - nrow(correct), "\n")

# Side-by-side visualization
par(mfrow = c(1, 2))
graphviz.plot(true_dag, main = "True DAG (what we built)",
              shape = "ellipse",
              highlight = list(nodes = "NMR", fill = "#E24A3F",
                               col = "#E24A3F"))
graphviz.plot(learned_dag, main = "Learned DAG (what hc() found)",
              shape = "ellipse",
              highlight = list(nodes = "NMR", fill = "#E24A3F",
                               col = "#E24A3F"))
```

### Adding Expert Knowledge as Constraints

The algorithm might get some arrows backwards or miss edges entirely. We can help it by providing "whitelisted" edges (arrows we know must exist) and "blacklisted" edges (arrows we know cannot exist). This is where expert knowledge enters the picture.

```{r}
#| label: expert-constraints
#| echo: true

# Expert says: Training MUST cause Staffing (not the reverse)
# Expert says: NMR CANNOT cause DrugAvail (outcomes don't cause inputs)
whitelist <- data.frame(
  from = c("Training"),
  to   = c("Staffing")
)
blacklist <- data.frame(
  from = c("NMR", "NMR", "NMR", "CSection", "Referral", "Volume"),
  to   = c("DrugAvail", "Training", "Equipment",
           "Training", "Training", "Training")
)

# Re-run with expert constraints
learned_expert <- hc(tz_data, score = "bic-cg",
                     whitelist = whitelist,
                     blacklist = blacklist)

cat("SHD without expert help:", shd(learned_dag, true_dag), "\n")
cat("SHD with expert help:   ", shd(learned_expert, true_dag), "\n")
cat("(Lower = closer to the truth)\n")
```

## Key Takeaways

::: {.key-takeaway}
**Structure learning algorithms** can discover candidate causal structures from data, but they cannot resolve all ambiguities. Multiple DAGs may be equally consistent with the observed data (Markov equivalence).

**The hill-climbing algorithm** (`bnlearn::hc()`) is a practical tool for exploring what the data suggests. It is fast, handles mixed data types, and gives you a starting point for expert review.

**Expert knowledge + algorithms = best results.** Use whitelists and blacklists to encode what you know for certain, and let the algorithm figure out the rest.

**The feedback loop** -- data, algorithm, expert review, decision, new data -- is how real health programs get smarter over time. No model is ever finished; it keeps improving as new evidence comes in.
:::

## Course Summary

Congratulations -- you have made it through the entire **Inference and Intervention** course. Here is what you have learned:

| Chapters | What You Learned |
|----------|-----------------|
| 1--3 | How to draw causal diagrams (DAGs) and read them like a language |
| 4--5 | How to turn diagrams into quantitative Bayesian networks and run diagnostic inference |
| 6 | How aggregated data can trick you (Simpson's Paradox) and how to avoid reasoning traps |
| 7 | How to make optimal decisions under uncertainty using influence diagrams and expected value |
| 8 | How to allocate resources across multiple countries using Monte Carlo simulation |
| 9 | How to anticipate strategic behavior using game theory and design better contracts |
| **10** | **How to let data help discover causal structures, guided by expert knowledge** |

Causal models are not academic exercises. They are practical tools for answering the questions that matter most: Where should we invest? What should we do? How do we know it will work? Whether you end up working in public health, business, policy, or any other field, this way of thinking -- making your assumptions explicit, testing them against evidence, and updating when you learn something new -- will serve you well.
