---
title: "Chapter 10 — Can the Computer Just Tell Us?"
subtitle: "Esther, the DHIS2 database, and the wall"
---

## The mystery

A data scientist at the Tanzania Ministry of Health sits in front of ten years of monthly facility-level data — 60,000 rows, 40 columns. Her director walks in and says *"just run an algorithm and tell us what's causing what"*. Modern algorithms can do remarkable things. But there is one thing they fundamentally cannot do — and the *cannot* is the most important lesson in the course.

::: {.callout-note}
**The one thing to remember from this chapter:** Data alone cannot tell you the direction of a causal arrow. Data plus a few expert assumptions can — and that is what structure learning really is.
:::

## Watch the lecture

::: {.slide-embed}
<iframe src="../ch10-structure-learning/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch10-structure-learning/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R workshop — run the PC algorithm, then run into the wall

In the lecture, Esther handed her data to a structure-learning algorithm and got back a network with mostly *undirected* edges. In this workshop you'll do the same on a small simulated MNH dataset and watch the wall appear in your console.

### Setting up

```{r}
#| label: ch10-setup
library(bnlearn)
library(dplyr)

set.seed(42)
```

### Build a small simulated dataset

We'll generate data from a *known* causal structure, then ask the algorithm to recover it. The known truth is:

- `RegionalBudget` → `Stockouts`
- `RegionalBudget` → `StaffShortage`
- `Stockouts` → `MaternalDeaths`
- `StaffShortage` → `MaternalDeaths`

```{r}
#| label: ch10-data
n <- 1000

regional_budget <- sample(c("low", "high"), n, replace = TRUE, prob = c(0.5, 0.5))

stockouts <- ifelse(regional_budget == "low",
                    sample(c("yes", "no"), n, replace = TRUE, prob = c(0.7, 0.3)),
                    sample(c("yes", "no"), n, replace = TRUE, prob = c(0.2, 0.8)))

staff_shortage <- ifelse(regional_budget == "low",
                         sample(c("yes", "no"), n, replace = TRUE, prob = c(0.6, 0.4)),
                         sample(c("yes", "no"), n, replace = TRUE, prob = c(0.2, 0.8)))

maternal_deaths <- ifelse(stockouts == "yes" | staff_shortage == "yes",
                          sample(c("high", "low"), n, replace = TRUE, prob = c(0.6, 0.4)),
                          sample(c("high", "low"), n, replace = TRUE, prob = c(0.15, 0.85)))

dhis_data <- data.frame(
  RegionalBudget  = factor(regional_budget),
  Stockouts       = factor(stockouts),
  StaffShortage   = factor(staff_shortage),
  MaternalDeaths  = factor(maternal_deaths)
)

head(dhis_data)
```

### Run the PC algorithm

```{r}
#| label: ch10-pc
# The PC algorithm — Spirtes-Glymour-Scheines (2000), the foundational
# structure-learning algorithm.
learned_network <- pc.stable(dhis_data)

learned_network
```

### Look at the discovered structure

```{r}
#| label: ch10-plot
plot(learned_network)

# What edges did the algorithm find?
arcs(learned_network)

# Which of those edges are DIRECTED, and which are still undirected?
directed.arcs(learned_network)
undirected.arcs(learned_network)
```

What you should see: the algorithm found roughly the right *connectivity* — it identified that `RegionalBudget` is linked to both `Stockouts` and `StaffShortage`, and that both of those are linked to `MaternalDeaths`. But many of the edges came back **undirected**. That's the wall from the lecture.

Even though we *built the data* from a clear causal structure, the algorithm could not always recover the directions from the data alone — because multiple causal stories produce the same conditional-independence pattern.

### Now hand the algorithm an expert hint

```{r}
#| label: ch10-expert
# Esther tells the algorithm: "I'm absolutely certain that RegionalBudget
# is upstream of everything else — it's a budget cycle, it can't be
# caused by stockouts." That's a domain assumption the algorithm cannot
# discover on its own.
blacklist <- data.frame(
  from = c("Stockouts", "StaffShortage", "MaternalDeaths",
           "Stockouts", "StaffShortage", "MaternalDeaths"),
  to   = c("RegionalBudget", "RegionalBudget", "RegionalBudget",
           "RegionalBudget", "RegionalBudget", "RegionalBudget")
)

learned_with_expert <- pc.stable(dhis_data, blacklist = blacklist)

plot(learned_with_expert)
arcs(learned_with_expert)
```

What you should see: with the single domain hint *"nothing causes RegionalBudget"*, more edges become directed. The algorithm could not discover this on its own, but once an expert gives it the orientation, the rest of the picture falls into place.

**That is the loop the lecture describes.** Algorithm + expert > algorithm alone > expert alone.

## Try it yourself

You run a structure-learning algorithm on a Kenyan county dataset. It finds a strong undirected edge between *facility births* and *newborn deaths*. What are at least three different stories the data could be telling? Hint: review chains, forks, colliders, and hidden variables — and don't forget the most dangerous explanation: that the algorithm picked up a reporting artefact, not a real causal link.

## You finished the course

You started in Chapter 1 with Almaz in Tigray and one question — *did the Health Extension Worker save her baby?* You now have the toolkit to answer that question, properly, the next time you encounter it in your own work. **Pick one programme you actually work on. Sit down with a pen. Draw the picture. Go interview three people. That single exercise will change how you think about your work — not in five years, but tomorrow.**
